---
title: "final_report"
author: "Youn Kyeong Chang (uni# yc3242)"
date: "December 15, 2018"
output:
  pdf_document: default
  github_document: default
---

#Final Project: Smoking in the Framingham Heart Study


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(mice)
library(gee)
library(lme4)
library(RLRsim)

frmgham = 
  read.csv2("frmgham2.csv", sep = ",", na.strings = c("","NA")) %>% 
  clean_names() %>% select(cursmoke, everything()) %>% 
  mutate(sex = as.factor(sex),
         bpmeds = as.factor(bpmeds),
         educ = as.factor(educ),
         diabetes = as.factor(diabetes),
         prevap = as.factor(prevap),
         prevchd = as.factor(prevchd),
         prevmi = as.factor(prevmi),
         prevstrk = as.factor(prevstrk),
         prevhyp = as.factor(prevhyp),
         sysbp = as.numeric(levels(sysbp))[sysbp],
         diabp = as.numeric(as.character(diabp)),
         bmi = as.numeric(as.character(bmi)),
         sex = if_else(sex == 1, "male", "female"),
         sex = fct_relevel(sex, "male")) %>% 
  as.tibble()
```

```{r}
frmgham = 
  frmgham %>% 
  mutate(hypertension
         = case_when(
      c(sysbp < 120 & diabp < 80) ~ "normal",  
      c((sysbp >= 120 & sysbp < 130) & diabp < 80) ~ "elevated",
      c((sysbp >= 130 & sysbp < 140) | (diabp >= 80 & diabp < 90)) ~ "stage1",
      c((sysbp >= 140 & sysbp < 180) | (diabp >= 90 & diabp < 120)) ~ "stage2",
      c((sysbp >= 180) | (diabp > 120)) ~ "crisis"),
      hypertension = fct_relevel(hypertension, "normal"))
```

# Descriptive statistics
```{r}
#descriptive statistics for the data

summary(frmgham)
```


#missing values
```{r}
frmgham %>%
  group_by(period) %>%
  summarize(sum_na_period = sum(is.na(period)))

d1 = frmgham %>%
  group_by(randid) %>%
  summarize(n = n())
table(d1$n)
# 447 patients have only one obs
# 781 patients have only two obs
# 3206 patients have all three obs

number.subjects <-
  frmgham  %>%
  select(randid) %>%
  distinct() %>%
  unlist() %>%
  length()

number.subjects

# missing data pattern by variables

sort(sapply(frmgham, function(x) { sum(is.na(x)) }), decreasing=TRUE)

```


* The output tells us that 2243 samples are complete, 7077 samples miss both hdlc and ldlc, 4 samples miss only the glucose value and so on.


* there are 447 patients who have only one obs; there are 781 patients who have only two obs and 3206 patients have all three obs. In total, there are 4434 patients in the study.

##1. Target question:
1) Is there a relationship between age and smoking status?  Does this relationship differ by sex? 

### missing values
```{r}
frmgham %>%
  group_by(period) %>%
  summarize(missing = number.subjects - n(), 
            sum_na_cursmoke = sum(is.na(cursmoke)))
```

### cursmoke vs age_ctg without adjustment
```{r}
mean(frmgham$age)
sd(frmgham$age)
min(frmgham$age)
max(frmgham$age)

## created age category variable (age_ctg)
frmgham =
  frmgham %>% 
  mutate(
      age_ctg = case_when(
      age < 35 ~ "35-",  
      age %in% 35:44 ~ "35-44",
      age %in% 45:54 ~ "45-54",
      age %in% 55:64 ~ "55-64",
      age %in% 65:74 ~ "65-74",
      age >= 75 ~ "75+"
    ),
    age_ctg = fct_relevel(age_ctg, "35-")) 

### cursmoke vs age_ctg : associated without adjustment 
frmgham %>% 
  select(randid, cursmoke, age_ctg) %>% 
  na.omit() %>% 
  group_by(age_ctg) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n()) %>% 
  ggplot(aes(x = age_ctg, y = p_cursmoke)) + 
  geom_bar(stat = "identity") +
  labs(title = "Relationship between categorized age and smoking status",
       x = "Age",
       y = "Probability of current smoking")

### cursmoke vs age_ctg : associated without adjustment 
summary_cursmoke = 
  frmgham %>% 
  select(randid, cursmoke, age_ctg) %>% 
  na.omit() %>% 
  group_by(age_ctg) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n()) 

mean(summary_cursmoke$p_cursmoke)
sd(summary_cursmoke$p_cursmoke)
min(summary_cursmoke$p_cursmoke)
max(summary_cursmoke$p_cursmoke)
```


## Finding modifier

```{r}
### by sex -- modified!
frmgham %>% 
  mutate(sex = ifelse(sex == 1, "male", "female")) %>%
  select(randid, cursmoke, age_ctg, sex) %>% 
  na.omit() %>% 
  group_by(age_ctg, sex) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n(),
            lower_CI = prop.test(sum(cursmoke), n())$conf.int[1],
            upper_CI = prop.test(sum(cursmoke), n())$conf.int[2]) %>%
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = sex, group = sex)) +
  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI), width = .1) +
  geom_line() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")

### by educ -- not modified except 35-

frmgham %>% 
  select(randid, cursmoke, age_ctg, educ) %>% 
  na.omit() %>% 
  group_by(age_ctg, educ) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n()) %>% 
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = educ, group = educ)) + 
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")

### by diabetes -- not modified 

frmgham %>% 
  select(randid, cursmoke, age_ctg, diabetes) %>% 
  na.omit() %>% 
  group_by(age_ctg, diabetes) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n()) %>% 
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = diabetes, group = diabetes)) + 
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")

### by prevap -- not modified 

frmgham %>% 
  select(randid, cursmoke, age_ctg, prevap) %>% 
  na.omit() %>% 
  group_by(age_ctg, prevap) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n()) %>% 
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = prevap, group = prevap)) + 
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")

### by prevchd -- not modified 

frmgham %>% 
  select(randid, cursmoke, age_ctg, prevchd) %>% 
  na.omit() %>% 
  group_by(age_ctg, prevchd) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n()) %>% 
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = prevchd, group = prevchd)) + 
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")

### by prevmi -- not modified 

frmgham %>% 
  select(randid, cursmoke, age_ctg, prevmi) %>% 
  na.omit() %>% 
  group_by(age_ctg, prevmi) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n(),
            lower_CI = prop.test(sum(cursmoke), n())$conf.int[1],
            upper_CI = prop.test(sum(cursmoke), n())$conf.int[2]) %>%
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = prevmi, group = prevmi)) +
  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI), width = .1) +
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")

### by prevstrk -- not modified 

frmgham %>% 
  select(randid, cursmoke, age_ctg, prevstrk) %>% 
  na.omit() %>% 
  group_by(age_ctg, prevstrk) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n(),
            lower_CI = prop.test(sum(cursmoke), n())$conf.int[1],
            upper_CI = prop.test(sum(cursmoke), n())$conf.int[2]) %>%
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = prevstrk, group = prevstrk)) +
  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI), width = .1) +
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")

### by prevhyp -- slightly modified 

frmgham %>% 
  select(randid, cursmoke, age_ctg, prevhyp) %>% 
  na.omit() %>% 
  group_by(age_ctg, prevhyp) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n(),
            lower_CI = prop.test(sum(cursmoke), n())$conf.int[1],
            upper_CI = prop.test(sum(cursmoke), n())$conf.int[2]) %>%
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = prevhyp, group = prevhyp)) +
  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI), width = .1) + 
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")

### by prevhyp -- slightly modified 

frmgham %>% 
  select(randid, cursmoke, age_ctg, hypertension) %>% 
  na.omit() %>% 
  group_by(age_ctg, hypertension) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n(),
            lower_CI = prop.test(sum(cursmoke), n())$conf.int[1],
            upper_CI = prop.test(sum(cursmoke), n())$conf.int[2]) %>%
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = hypertension, group = hypertension)) +
  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI), width = .1) + 
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")
```

## Finding confounders 
### age and cursmoke -- confounder: hdlc, ldlc

```{r}
age <-
  glmer(cursmoke ~ age + (1|randid),
      data = frmgham,
      family = binomial, 
      nAGQ = 0)

pot_conf <- names(frmgham)[-c(1:3, 5)]
pot_conf <-
  pot_conf[-c(grep("time", pot_conf))]

conf_det <- vector("list", length = length(pot_conf))
names(conf_det) <- pot_conf



for (conf in pot_conf) {

  data <- cbind(frmgham[,c("cursmoke", "age", "randid")], frmgham[,conf])
  formula <- paste("cursmoke","~","age + (1|randid) +", conf)
  try(fit_glmer <- glmer(formula,
      data = data,
      family = binomial,
      nAGQ = 0))

  result <- tibble(
    variable = c("age"),
    OR =
      exp(coef(summary(age))[2,1]),
    lower_CI =
      exp(coef(summary(age))[2,1] - 1.96*(coef(summary(age))[2,2])),
    upper_CI =
      exp(coef(summary(age))[2,1] + 1.96*(coef(summary(age))[2,2])),
    potconf_OR =
      exp(coef(summary(fit_glmer))[2,1]),
    confounder =
      if_else((potconf_OR >= lower_CI & potconf_OR <= upper_CI), FALSE, TRUE)
  )

  conf_det[[conf]] <- result
}

map(conf_det, knitr::kable, digits = 3)
```


## Fit model using glmer
```{r}
#educ variable
##https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4939617/

## From eda  
cursmoke_age <-
  glmer(cursmoke ~ age*sex + educ + hdlc + ldlc + age*prevhyp + (1|randid), 
        data = frmgham, family = binomial, nAGQ = 0)

summary(cursmoke_age) ## AIC 3581.6, var(random) = 0.007

# changed model based on cigpday result 
cursmoke_age <-
  glmer(cursmoke ~ age*sex + educ + heartrte + hdlc + ldlc + age*prevhyp + (1|randid), 
        data = frmgham, family = binomial, nAGQ = 0)

summary(cursmoke_age) ## AIC 3550.2, var(random) = 0.004

## of interst -- bpmeds 
cursmoke_age <-
  glmer(cursmoke ~ age*sex + educ + bpmeds + heartrte + hdlc + ldlc + age*prevhyp + (1|randid), 
        data = frmgham, family = binomial, nAGQ = 0)

summary(cursmoke_age) # AIC 3033.0, var = 0.004238

## of interst -- bpmeds + bmi
cursmoke_age <-
  glmer(cursmoke ~ age*sex + educ + bpmeds + bmi + heartrte + hdlc + ldlc + age*prevhyp + (1|randid),
        data = frmgham, family = binomial, nAGQ = 0)

summary(cursmoke_age) # AIC  2961.3, var = 4.151e-14

## of interst -- bpmeds + bmi + diabetes 
cursmoke_age <-
  glmer(cursmoke ~ age*sex + educ + bpmeds + heartrte + diabetes + hdlc + ldlc + age*prevhyp + (1|randid), 
        data = frmgham, family = binomial, nAGQ = 0)

summary(cursmoke_age) # AIC  3034.9, var = 0.00417 

## of interst 
cursmoke_age <-
  glmer(cursmoke ~ age*sex + educ + bmi + heartrte + hypertension + hdlc + ldlc +  (1|randid), 
        data = frmgham, family = binomial, nAGQ = 0)
summary(cursmoke_age) # AIC  3466.1  , var = 0  

## of interst -- final----- 
cursmoke_age <-
  glmer(cursmoke ~ age*sex + educ + bpmeds + bmi + heartrte + hdlc + ldlc +  (1|randid), 
        data = frmgham, family = binomial, nAGQ = 0)
summary(cursmoke_age) # AIC  2964.4 , var = 4.024e-12 
exp(coef(summary(cursmoke_age)))[,1]
vcov(summary(cursmoke_age))
#p-value for age + age:sex2
pnorm((-0.0524146+ -0.0314291)/sqrt(6.965556e-05 + 1.323812e-04 + 2*(-6.762571e-05)))

## of interst -- hypertension + bmi + heartrte
cursmoke_age <-
  glmer(cursmoke ~ age*sex + educ + hypertension + bmi + heartrte + hdlc + ldlc + (1|randid), 
        data = frmgham, family = binomial, nAGQ = 0)

summary(cursmoke_age) # AIC  3466.1  , var =  4e-14 

## of interst -- hypertension + bmi + heartrte
cursmoke_age <-
  glmer(cursmoke ~ age*sex + educ + hypertension + bmi + heartrte + hdlc + ldlc + (1|randid), 
        data = frmgham, family = binomial, nAGQ = 0)

summary(cursmoke_age) # AIC  3466.4   , var =  0 

# removed interaction of age*prevhyp
cursmoke_age <-
  glmer(cursmoke ~ age*sex + educ + heartrte + hdlc + ldlc + (1|randid), 
        data = frmgham, family = binomial, nAGQ = 0)

summary(cursmoke_age) ## AIC 3566.8, var(random) = 0.004

cursmoke_age <-
  glmer(cursmoke ~ age*sex + educ + heartrte + prevhyp + (1|randid),
        data = frmgham, family = binomial, nAGQ = 0)

summary(cursmoke_age) ## AIC  10648.8  var(random) = 21.03

```


(2) Is there a relationship between the number of cigarettes smoked per day and age?  Does this relationship differ by sex? 

While answering these questions, please account for any confounders that you have evidence may impact the relationship between age and sex with smoking.

```{r}
frmgham %>% 
  ggplot(., aes(x=age, y=cigpday)) +
    geom_jitter() + 
    geom_smooth(method = "loess") +
  ggtitle("Unadjusted relationship between the number of cigarettes smoked per day and age")
```

```{r}
frmgham_smoker  = frmgham %>%
filter(cigpday != 0) 

#summary stat 
frmgham_smoker %>%
  dplyr::select(cigpday, age, period) %>%
  pastecs::stat.desc() %>%
  round(digits = 3)

frmgham_smoker %>%
  distinct(randid) %>%
  nrow()
#2294 # of smokers

#missings in smokers
frmgham_smoker %>%
 group_by(period) %>%
   summarize(n = n(),
     missing = 2294 - n,
           sum_na_cigpday = sum(is.na(cigpday)))

frmgham_smoker %>%
 group_by(randid) %>%
 mutate(mean.cigpday = mean(cigpday)) %>%
 ungroup() %>%
 select(randid, mean.cigpday) %>%
 distinct() %>%
 ggplot(aes(x = mean.cigpday)) +
 geom_histogram(color="darkblue", fill="lightblue", aes(y=..density..)) +
 labs(x = "Average number of cigarettes smoked per day",
      y = "Density",
      title = "cigpday summary")


```


```{r spaghetti plot}
frmgham_smoker %>% 
  mutate(sex = ifelse(sex == 1, "male", "women")) %>%
  ggplot(aes(x = age, y = cigpday, group = randid, color = sex)) +
  geom_line(alpha = .5) 
```

```{r finding counfounders for m2}
# I will look into SYSBP DIABP TOTCHOL BMI DIABETES 
# based on https://tobaccocontrol.bmj.com/content/14/5/315

age <- glmer(cigpday ~ age + (1 | randid), 
      data = frmgham,
      family = poisson,
      nAGQ = 0)

pot_conf <- names(frmgham)[-c(2, 5, 8)]
pot_conf <-
  pot_conf[-c(grep("time", pot_conf))]

conf_det <- vector("list", length = length(pot_conf))
names(conf_det) <- pot_conf

conf_det <- vector("list", length = length(pot_conf))
names(conf_det) <- pot_conf

for (conf in pot_conf) {

  data <- cbind(frmgham[,c("cigpday", "age", "randid")], frmgham[,conf])
  formula <- paste("cigpday","~","age + (1|randid) +", conf)
  
  try(fit_glmer <- glmer(formula,
      data = data,
      family = poisson,
      nAGQ = 0))

  result <- tibble(
    variable = c("age"),
    OR =
      exp(coef(summary(age))[2,1]),
    lower_CI  =
      exp(coef(summary(age))[2,1] - 1.96*(coef(summary(age))[2,2])),
    upper_CI =
      exp(coef(summary(age))[2,1] + 1.96*(coef(summary(age))[2,2])),
    potconf_OR =
      exp(coef(summary(fit_glmer))[2,1]),
    confounder =
      if_else((potconf_OR >= lower_CI & potconf_OR <= upper_CI), FALSE, TRUE)
  )

  conf_det[[conf]] <- result
}

#map(conf_det, knitr::kable, digits = 3)
conf_det
```

```{r}
frmgham_smoker %>% 
  mutate(sex = ifelse(sex == 1, "male", "women")) %>%
  select(randid, cigpday, age_ctg, sex) %>% 
  na.omit() %>% 
  group_by(age_ctg, sex) %>% 
  summarise(cigpday_mean = mean(cigpday),
            se = sqrt(var(cigpday)/n()),
            n = n()) %>%
  ggplot(aes(x = age_ctg, y = cigpday_mean, color = sex, group = sex)) + 
  geom_line() +
  geom_errorbar(aes(ymin = cigpday_mean - se, ymax = cigpday_mean + se), width = .1) + 
  labs(title = "Relationship between categorized age and cigpday by sex",
       x = "Age",
       y = "mean number of cigarettes smoked per day")


```


```{r backward selection}
cigpday.lmer1<- glmer(cigpday ~ age + sex + heartrte + prevchd + hdlc + prevap + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer1)
# AIC   7873.592 
cigpday.lmer2<- glmer(cigpday ~ age + sex + heartrte + prevchd + hdlc + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer2)
# AIC 7874.8
# remove prevchd
cigpday.lmer3<- glmer(cigpday ~ age + sex + heartrte + hdlc + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer3)
# AIC 7872.8
#-0.002262
cigpday.lmer3<- glmer(cigpday ~ age + sex + heartrte + hdlc + prevchd + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer3)

# tested interaction term with age
cigpday.lmer4<- glmer(cigpday ~ age + sex + heartrte + hdlc + age:sex + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer4)
# AIC  7873.8
cigpday.lmer5<- glmer(cigpday ~ age + sex + heartrte + hdlc + age:heartrte + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer5)
# decided with heartrte
# AIC   7870.3
```



```{r model selection based on research}
cigpday.lmer6<- glmer(cigpday ~ age + sex + educ + heartrte + prevhyp + age*sex + age:prevhyp + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer6)
# AIC 35492.2

cigpday.lmer7<- glmer(cigpday ~ age + sex + educ + heartrte + prevhyp + hdlc + age:sex + age:prevhyp + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer7)
# AIC 7678.7
# need to include hdlc

##** FINAL MODEL **##
# removing prevhyp and age:prevhyp add age:heartrte
cigpday.lmer8<- glmer(cigpday ~ age + sex + educ + heartrte + hdlc + age:heartrte + age:sex + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer8)
# AIC 7672.4 

##** FINAL MODEL **##
# removing educ
cigpday.lmer9<- glmer(cigpday ~ age + sex + heartrte + hdlc + age:heartrte + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer9)
# AIC 7870.3

exp(0.0190057)
```
the var for random intercept; randid (Intercept) 0.3513 is low
test overspersion

```{r model assumption checking}
# randef qqplot
qqnorm(ranef(cigpday.lmer9)$randid[,1])
qqline(ranef(cigpday.lmer9)$randid[,1])

plot(fitted(cigpday.lmer9), resid(cigpday.lmer9)) #residuals vs fitted
abline(h=0)

smoker_rmna <- frmgham_smoker %>%
  dplyr::select(randid, cigpday, age, sex, heartrte, hdlc) %>% drop_na()
smoker_rmna$fitted <- fitted(cigpday.lmer9)    

#fitted vs observed
plot(smoker_rmna$fitted, jitter(smoker_rmna$cigpday,0.1))
abline(0,1)

# overdispersion func 
# https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#overdispersion
overdisp_fun <- function(model) {
    rdf <- df.residual(model)
    rp <- residuals(model,type="pearson")
    Pearson.chisq <- sum(rp^2)
    prat <- Pearson.chisq/rdf
    pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
    c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}
overdisp_fun(cigpday.lmer9)
# tested the goodness of fit using Chi square test 
# null hypothesis is that our model is correctly specified
# p-val = 1 not reject, no overdispersion

####testing overdispersion on glm REMOVE???#####
#GLM for single time point 
fit.glm <- frmgham_smoker %>%
  group_by(randid) %>%
  summarise(min_age = min(age)) %>%
  left_join(frmgham_smoker, by = c("min_age" = "age", "randid" = "randid")) %>%
  glm(cigpday ~ min_age + sex + heartrte + hdlc + min_age*heartrte,
                   data = .,
                   family = 'poisson')
summary(fit.glm)
# Null deviance: 100.78  on 15  degrees of freedom
# overdispersion
```



#2.Target question:

Next you are interested in the relationship between certain health outcomes and smoking status. In particular you are interested in :


(1)The relationship between current smoking status and systolic blood pressure.


```{r}
#jitter plot of cursmoke and sysbp
frmgham %>% 
  ggplot(., aes(x=cursmoke, y=sysbp)) +
  geom_jitter() + 
  geom_smooth(method = "loess") + 
  ggtitle("Unadjusted relationship between current smoking status and systolic blood pressure")

frmgham = frmgham %>% mutate(cursmoke = as.factor(cursmoke),
                             period = as.factor(period))

##check random effects
ggplot(data = frmgham, aes(x = period, y = sysbp, group = randid)) +
  geom_point() +
  geom_line() +
  facet_wrap( ~ cursmoke, ncol=2) +
  ggtitle("sysbp versus period for smoker and nonsmoker")

syslm = lm(sysbp~cursmoke, data = frmgham)
sysbp <- lmer(sysbp ~ cursmoke + (1|randid), data = frmgham)
# exactLRT(sysbp, syslm)
##p-value < 2.2e-16, random effect is significant.

#find the confounder
sysbp_sex <- lmer(sysbp ~ cursmoke + sex + (1|randid), data = frmgham)
sysbp_totchol <- lmer(sysbp ~ cursmoke + totchol + (1|randid), data = frmgham)
sysbp_age <- lmer(sysbp ~ cursmoke + age + (1|randid), data = frmgham)
sysbp_diabp <- lmer(sysbp ~ cursmoke + diabp + (1|randid), data = frmgham)
sysbp_bmi <- lmer(sysbp ~ cursmoke + bmi + (1|randid), data = frmgham)
sysbp_diab <- lmer(sysbp ~ cursmoke + diabetes + (1|randid), data = frmgham)
sysbp_bpmeds <- lmer(sysbp ~ cursmoke + bpmeds + (1|randid), data = frmgham)
sysbp_heart <- lmer(sysbp ~ cursmoke + heartrte + (1|randid), data = frmgham)
sysbp_glu <- lmer(sysbp ~ cursmoke + glucose + (1|randid), data = frmgham)
sysbp_educ <- lmer(sysbp ~ cursmoke + educ + (1|randid), data = frmgham)
sysbp_prevchd <- lmer(sysbp ~ cursmoke + prevchd + (1|randid), data = frmgham)
sysbp_prevap <- lmer(sysbp ~ cursmoke + prevap + (1|randid), data = frmgham)
sysbp_prevmi <- lmer(sysbp ~ cursmoke + prevmi + (1|randid), data = frmgham)
sysbp_strk<- lmer(sysbp ~ cursmoke + prevstrk + (1|randid), data = frmgham)
sysbp_hyp <- lmer(sysbp ~ cursmoke + prevhyp + (1|randid), data = frmgham)
sysbp_per <- lmer(sysbp ~ cursmoke + period + (1|randid), data = frmgham)
tibble(
  variable = c("sex", "totchol", "age", "diabp", "bmi", "diab", "bpmeds",
               "heart", "glu", "educ", "prevchd", "prevap", "prevmi", "prevstrk", "prevhyp",
               "period"),
  OR = 
    exp(coef(summary(sysbp))[2,1]),
  lower_CI  = 
    exp(coef(summary(sysbp))[2,1] - 1.96*(coef(summary(sysbp))[2,2])),
  upper_CI = 
    exp(coef(summary(sysbp))[2,1] + 1.96*(coef(summary(sysbp))[2,2])),
  conf_OR =
    c(exp(coef(summary(sysbp_sex))[2,1]),
      exp(coef(summary(sysbp_totchol))[2,1]),
      exp(coef(summary(sysbp_age))[2,1]),
      exp(coef(summary(sysbp_diabp))[2,1]),
      exp(coef(summary(sysbp_bmi))[2,1]),
      exp(coef(summary(sysbp_diab))[2,1]),
      exp(coef(summary(sysbp_bpmeds))[2,1]),
      exp(coef(summary(sysbp_heart))[2,1]),
      exp(coef(summary(sysbp_glu))[2,1]),
      exp(coef(summary(sysbp_educ))[2,1]),
      exp(coef(summary(sysbp_prevchd))[2,1]),
      exp(coef(summary(sysbp_prevap))[2,1]),
      exp(coef(summary(sysbp_prevmi))[2,1]),
      exp(coef(summary(sysbp_strk))[2,1]),
      exp(coef(summary(sysbp_hyp))[2,1]),
      exp(coef(summary(sysbp_per))[2,1])),
  confounder =
    if_else((conf_OR >= lower_CI & conf_OR <= upper_CI), FALSE, TRUE)
)

#model adding the confounders
sysbp_smoke = lmer(sysbp ~ cursmoke+ age + diabp + bmi + prevhyp+ period + (1|randid), 
                    data = frmgham)

#check interaction
sysbp1 = lmer(sysbp ~ cursmoke*age + diabp + bmi + prevhyp+ period + (1|randid), 
              data = frmgham)
sysbp2 = lmer(sysbp ~ cursmoke*diabp + age + bmi + prevhyp +period+ (1|randid), 
              data = frmgham)
sysbp3 = lmer(sysbp ~ cursmoke*bmi + age + diabp + prevhyp+period + (1|randid), 
              data = frmgham)
sysbp4 = lmer(sysbp ~ cursmoke*prevhyp + age + bmi + diabp +period+ (1|randid), 
              data = frmgham)
sysbp5 = lmer(sysbp ~ cursmoke*period + age + bmi + diabp +prevhyp+ (1|randid), 
              data = frmgham)
summary(sysbp1)
summary(sysbp2)
summary(sysbp3)
summary(sysbp4)
summary(sysbp5)
AIC(sysbp1)
#interaction term reference: Ulrich John, Monika Hanke,Christian Meyer,Anja Schumann. 
#Gender and age differences among current smokers in a general population survey. BMC Public Health. 2005; 5: 57.
#Published online 2005 Jun 3. doi:  [10.1186/1471-2458-5-57]
AIC(sysbp2)
AIC(sysbp3)
AIC(sysbp4)
AIC(sysbp5)

#95% CI for beta of cursmoke and interaction term
b_par<-bootMer(x=sysbp1,FUN=fixef,nsim=200)
boot::boot.ci(b_par,type="basic",index=2)
b_par<-bootMer(x=sysbp1,FUN=fixef,nsim=200)
boot::boot.ci(b_par,type="basic",index=9)
```


(2)The relationship between current smoking status and diastolic blood pressure.
```{r}
#jitter plot of cursmoke and diabp
frmgham %>% 
  ggplot(., aes(x=cursmoke, y=diabp)) +
  geom_jitter() + 
  geom_smooth(method = "loess") + 
  ggtitle("Unadjusted relationship between current smoking status and diastolic blood pressure")

#check random effects
ggplot(data = frmgham, aes(x = period, y = diabp, group = randid)) +
  geom_point() +
  geom_line() +
  facet_wrap( ~ cursmoke, ncol=2) +
  ggtitle("diabp versus period for smoker and nonsmoker")

dialm = lm(diabp~cursmoke, data = frmgham)
diabp <- lmer(diabp ~ cursmoke + (1|randid), data = frmgham)
# exactLRT(diabp, dialm)
#p-value < 2.2e-16, random effect is significant.

#find confounder
diabp_sex <- lmer(diabp ~ cursmoke + sex + (1|randid), data = frmgham)
diabp_totchol <- lmer(diabp ~ cursmoke + totchol + (1|randid), data = frmgham)
diabp_age <- lmer(diabp ~ cursmoke + age + (1|randid), data = frmgham)
diabp_sysbp <- lmer(diabp ~ cursmoke + sysbp + (1|randid), data = frmgham)
diabp_bmi <- lmer(diabp ~ cursmoke + bmi + (1|randid), data = frmgham)
diabp_diab <- lmer(diabp ~ cursmoke + diabetes + (1|randid), data = frmgham)
diabp_bpmeds <- lmer(diabp ~ cursmoke + bpmeds + (1|randid), data = frmgham)
diabp_heart <- lmer(diabp ~ cursmoke + heartrte + (1|randid), data = frmgham)
diabp_glu <- lmer(diabp ~ cursmoke + glucose + (1|randid), data = frmgham)
diabp_educ <- lmer(diabp ~ cursmoke + educ + (1|randid), data = frmgham)
diabp_prevchd <- lmer(diabp ~ cursmoke + prevchd + (1|randid), data = frmgham)
diabp_prevap <- lmer(diabp ~ cursmoke + prevap + (1|randid), data = frmgham)
diabp_prevmi <- lmer(diabp ~ cursmoke + prevmi + (1|randid), data = frmgham)
diabp_strk<- lmer(diabp ~ cursmoke + prevstrk + (1|randid), data = frmgham)
diabp_hyp <- lmer(diabp ~ cursmoke + prevhyp + (1|randid), data = frmgham)
diabp_per <- lmer(diabp ~ cursmoke + period + (1|randid), data = frmgham)
tibble(
  variable = c("sex", "totchol", "age", "sysbp", "bmi", "diab", "bpmeds",
               "heart", "glu", "educ", "prevchd", "prevap", "prevmi", "prevstrk", "prevhyp",
               "period"),
  OR = 
    exp(coef(summary(diabp))[2,1]),
  lower_CI  = 
    exp(coef(summary(diabp))[2,1] - 1.96*(coef(summary(diabp))[2,2])),
  upper_CI = 
    exp(coef(summary(diabp))[2,1] + 1.96*(coef(summary(diabp))[2,2])),
  conf_OR =
    c(exp(coef(summary(diabp_sex))[2,1]),
      exp(coef(summary(diabp_totchol))[2,1]),
      exp(coef(summary(diabp_age))[2,1]),
      exp(coef(summary(diabp_sysbp))[2,1]),
      exp(coef(summary(diabp_bmi))[2,1]),
      exp(coef(summary(diabp_diab))[2,1]),
      exp(coef(summary(diabp_bpmeds))[2,1]),
      exp(coef(summary(diabp_heart))[2,1]),
      exp(coef(summary(diabp_glu))[2,1]),
      exp(coef(summary(diabp_educ))[2,1]),
      exp(coef(summary(diabp_prevchd))[2,1]),
      exp(coef(summary(diabp_prevap))[2,1]),
      exp(coef(summary(diabp_prevmi))[2,1]),
      exp(coef(summary(diabp_strk))[2,1]),
      exp(coef(summary(diabp_hyp))[2,1]),
      exp(coef(summary(diabp_per))[2,1])),
  confounder =
    if_else((conf_OR >= lower_CI & conf_OR <= upper_CI), FALSE, TRUE)
)

#model adding the confounders
diabp_smoke = lmer(diabp ~ cursmoke + sysbp + bmi + prevhyp + (1|randid), 
                   data = frmgham)
summary(diabp_smoke)

#check interaction
diabp1 = lmer(diabp ~ cursmoke*sysbp + bmi + prevhyp + (1|randid), 
              data = frmgham)
diabp2 = lmer(diabp ~ cursmoke*bmi +sysbp+ prevhyp + (1|randid), 
              data = frmgham)
diabp3 = lmer(diabp ~ cursmoke*prevhyp + bmi + sysbp +(1|randid), 
              data = frmgham)
summary(diabp1)
summary(diabp2)
summary(diabp3)
AIC(diabp1)
AIC(diabp2)
AIC(diabp3)
AIC(diabp_smoke)

#95% CI of beta of cursmoke and interaction term
b_par<-bootMer(x=diabp1,FUN=fixef,nsim=200)
boot::boot.ci(b_par,type="basic",index=2)
b_par<-bootMer(x=diabp1,FUN=fixef,nsim=200)
boot::boot.ci(b_par,type="basic",index=6)
```



(3)The relationship between current smoking status and serum total cholesterol. Again, while answering these questions, please account for any confounders that you have evidence may impact these relationships. 




```{r}
frmgham %>% 
  ggplot(., aes(x=cursmoke, y=totchol)) +
    geom_bar(stat="identity") +ggtitle("Unadjusted relationship between current smoking status and serum total cholesterol")


frmgham %>% 
  ggplot(., aes(x=as.factor(cursmoke), y=totchol)) +
  geom_boxplot() +ggtitle("Serum total cholesterol by current smoking status, Boxplot")


frmgham %>% 
  ggplot(., aes(x=as.factor(cursmoke), y=totchol)) +
  geom_violin(aes(fill = as.factor(cursmoke))) +ggtitle("Serum total cholesterol by current smoking status, Violin")
#https://medlineplus.gov/cholesterollevelswhatyouneedtoknow.html
#categorized totchol based on link above

frmgham=frmgham %>% 
  mutate(totchol_ctg= ifelse(age<=19 & totchol<170,0,
                             ifelse(age>19&totchol %in% c(125, 200), 0,1)))
table(frmgham$totchol_ctg)
# 124 normal total chol 
# 11503 abnormal total chol

```


## Finding confounders


```{r}

library(nlme)

totchol <- gls(totchol ~ as.factor(cursmoke), data = frmgham,na.action=na.omit)
totchol3=lmer(totchol ~ as.factor(cursmoke)+age+sex+prevchd+(1|randid),
              data=frmgham)

totchol4=lmer(totchol ~ as.factor(cursmoke)*age+sex+prevchd+(as.factor(cursmoke)|randid),
              data=frmgham)

totchol5=lmer(totchol ~ as.factor(cursmoke)+age+sex+prevchd+bmi+(1|randid),
              data=frmgham)
totchol6=lmer(totchol ~ as.factor(cursmoke)+age+sex+prevchd+bmi+(1|randid),
              data=frmgham)

frmgham=frmgham %>% 
  mutate(period = period-1)

totchol7=lmer(totchol ~ as.factor(cursmoke)*sex+bmi+age+prevchd+(1|randid),
              data=frmgham)
totchol8=lmer(totchol ~ as.factor(cursmoke)*sex+bmi+age+(1|randid),
              data=frmgham)
totchol9=lmer(totchol ~ as.factor(cursmoke)+sex+bmi+age+(1|randid),
              data=frmgham)
# totchol11=lmer(totchol ~ as.factor(cursmoke)*as.factor(period)+sex+bmi+age+(1|randid),
#               data=frmgham)
# totchol10=lmer(totchol ~ as.factor(cursmoke)+as.factor(period)+sex+bmi+age+(1|randid),
#               data=frmgham)
# 
# totchol12=lmer(totchol ~ as.factor(cursmoke)+as.factor(period)+bmi+age+(1|randid),
#               data=frmgham)
# totchol13=lmer(totcholas.factor(cursmoke)*sex+bmi+age+prevstrk+prevchd+(1|randid),
#               data=frmgham)

#final model
totchol15=lmer(totchol ~ as.factor(cursmoke)*sex+bmi+age+sysbp+(1|randid),
              data=frmgham)
AIC(totchol15)


# AIC(totchol10)
# summary(totchol10)
# summary(totchol11)
# anova(totchol13,totchol14)
# test the significance of random effect
#sig_ranef=RLRsim::exactRLRT(totchol13, seed=1)

# pot_conf <- names(frmgham)[-c(1,2,4)]
# pot_conf <- pot_conf[-c(grep("time", pot_conf), grep("death", pot_conf))]
# 
# conf_det <- vector("list", length = length(pot_conf))
# names(conf_det) <- pot_conf
# 
# 
# 
# for (conf in pot_conf) {
# 
#  data <- cbind(frmgham[,c("totchol","cursmoke")], frmgham[,conf])
#  try(fit_gee <- gls(totchol ~ .,
#      data = data,
#      ,na.action=na.omit))
# 
#  result <- tibble(
#    variable = c("totchol"),
#    OR =
#      exp(coef(summary(totchol))[2,1]),
#    lower_CI  =
#      exp(coef(summary(totchol))[2,1] - 1.96*(coef(summary(totchol))[2,3])),
#    upper_CI =
#      exp(coef(summary(totchol))[2,1] + 1.96*(coef(summary(totchol))[2,3])),
#    potconf_OR =
#      exp(coef(summary(fit_gee))[2,1]),
#    confounder =
#      if_else((potconf_OR >= lower_CI & potconf_OR <= upper_CI), FALSE, TRUE)
#  )
# 
#  conf_det[[conf]] <- result
# }
# 
# map(conf_det, knitr::kable, digits = 3)
```


##visualize the relationship between totchol and cursmoke by each confounding variable:

```{r}

frmgham %>%
  mutate(cursmoke=as.factor(cursmoke)) %>% 
  group_by(cursmoke) %>%
  ggplot(.)+  
  geom_boxplot(aes(x=cursmoke, y=totchol)) +
  ggtitle("Totochol by smoking status, boxplot")


frmgham %>%
  mutate(cursmoke=as.factor(cursmoke)) %>% 
  group_by(cursmoke, sex) %>%
  summarise(
    n=n(),
    mean=mean(totchol, na.rm=T),
    sd=sd(totchol, na.rm=T)
  ) %>%
  mutate( se=sd/sqrt(n))  %>%
  mutate( ic=se * qt((1-0.05)/2 + .5, n-1)) %>%
  ggplot()+
  geom_bar(aes(x=cursmoke, y=mean), stat="identity", fill="forestgreen", alpha=0.5) +
  geom_errorbar( aes(x=cursmoke, ymin=mean-ic, ymax=mean+ic), width=0.4, colour="orange", alpha=0.9, size=0.75) +
  ggtitle("Mean totochol by gender")+facet_wrap(~sex)





```



## bootstrap for confidence interval


```{r}
# b_par<-bootMer(x=totchol15,FUN=fixef,nsim=200)
# a=boot::boot.ci(b_par,type="basic",index=2)
# a$basic[1,4]+b$basic[1,4]
# a$basic[1,5]+b$basic[1,5]
# 
# b=boot::boot.ci(b_par,type="basic",index=7)
# b
# fixef(totchol15)
```




### Figure 1 ##### 
```{r}
# library(patchwork)
# 
# figure1_1 = 
# frmgham %>% 
#   select(randid, cursmoke, age_ctg, sex) %>% 
#   na.omit() %>% 
#   group_by(age_ctg, sex) %>% 
#   summarise(p_cursmoke = sum(cursmoke)/n(),
#             lower_CI = prop.test(sum(cursmoke), n())$conf.int[1],
#             upper_CI = prop.test(sum(cursmoke), n())$conf.int[2]) %>%
#   ggplot(aes(x = age_ctg, y = p_cursmoke, color = sex, group = sex)) +
#   geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI), width = .1) +
#   geom_line(size = 1) +
#   labs(subtitle = "categorized age vs smoking status by sex",
#        x = "Age",
#        y = "Probability of current smoking") +
#   theme(plot.subtitle = element_text(size = 7),
#         axis.text.x = element_text(size = 7, angle = 45),
#         axis.text.y = element_text(size = 7),
#         axis.title.x = element_text(size = 8),
#         axis.title.y = element_text(size = 8),
#         legend.position = 'none') 
# 
# figure1_2 = 
#  frmgham_smoker %>%
#  select(randid, cigpday, age_ctg, sex) %>%
#  na.omit() %>%
#  group_by(age_ctg, sex) %>%
#  summarise(cigpday_mean = mean(cigpday),
#            se = sqrt(var(cigpday)/n()),
#            n = n()) %>%
#  ggplot(aes(x = age_ctg, y = cigpday_mean, group = sex, color = sex)) +
#  geom_line(size = 1) +
#  geom_errorbar(aes(ymin = cigpday_mean - se, ymax = cigpday_mean + se), width = .1) +
#  labs(subtitle = "categorized age vs cigpday by sex",
#       x = "Age",
#       y = "Mean number of cigarettes smoked per day") +
#    theme(plot.subtitle = element_text(size = 7),
#         axis.text.x = element_text(size = 7, angle = 45),
#         axis.text.y = element_text(size = 7),
#         axis.title.x = element_text(size = 8),
#         axis.title.y = element_text(size = 8),
#         legend.text = element_text(size = 5),
#         legend.title = element_text(size = 5))  
# 
# 
# figure1 = 
# figure1_1 + figure1_2 +
#   plot_annotation(title = "Figure 1. Relationship between the categorized age and the outcome by sex without adjustment", theme = theme(plot.title = element_text(size = 8)))

# ggsave("figure1.png", plot = figure1)
```

### Figure2 ###
```{r}
# fig2_1 = 
#   frmgham %>%
#  mutate(cursmoke=as.factor(cursmoke)) %>%
#  group_by(cursmoke) %>%
#  ggplot(.)+
#  geom_boxplot(aes(x=cursmoke, y=sysbp)) +
#  labs(subtitle = "systolic blood pressure") +
#  theme(plot.subtitle = element_text(size = 8),
#        axis.text.x = element_text(size = 7),
#        axis.text.y = element_text(size = 7),
#        axis.title.x = element_text(size = 8),
#        axis.title.y = element_text(size = 8))
#        
# fig2_2 = 
#   frmgham %>%
#  mutate(cursmoke=as.factor(cursmoke)) %>%
#  group_by(cursmoke) %>%
#  ggplot(.)+
#  geom_boxplot(aes(x=cursmoke, y=diabp)) +
#  labs(subtitle = "diastolic blood pressure") +
#  theme(plot.subtitle = element_text(size = 8),
#        axis.text.x = element_text(size = 7),
#        axis.text.y = element_text(size = 7),
#        axis.title.x = element_text(size = 8),
#        axis.title.y = element_text(size = 8))
# 
# fig2_3 = 
#   frmgham %>%
#  mutate(cursmoke=as.factor(cursmoke)) %>%
#  group_by(cursmoke) %>%
#  ggplot(.)+
#  geom_boxplot(aes(x=cursmoke, y=totchol)) +
#  labs(subtitle = "total cholesterol") +
#  theme(plot.subtitle = element_text(size = 8),
#        axis.text.x = element_text(size = 7),
#        axis.text.y = element_text(size = 7),
#        axis.title.x = element_text(size = 8),
#        axis.title.y = element_text(size = 8))
# 
# figure2 = 
# fig2_1 + fig2_2 + fig2_3 + plot_layout(ncol = 3) +
#   plot_annotation(title = "Figure 2. Relationship between current smoking stutus and the outcome without adjustment", theme = theme(plot.title = element_text(size = 8)))

# ggsave("figure2.png", plot = figure2)
```



