---
title: "final_report"
author: "Youn Kyeong Chang (uni# yc3242)"
date: "December 15, 2018"
output: github_document
---

#Final Project: Smoking in the Framingham Heart Study


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(janitor)
library(mice)
library(gee)
library(lme4)
library(RLRsim)

frmgham = 
  read.csv2("frmgham2.csv", sep = ",", na.strings = c("","NA")) %>% 
  clean_names() %>% select(cursmoke, everything()) %>% 
  mutate(sex = as.factor(sex),
         bpmeds = as.factor(bpmeds),
         educ = as.factor(educ),
         diabetes = as.factor(diabetes),
         prevap = as.factor(prevap),
         prevchd = as.factor(prevchd),
         prevmi = as.factor(prevmi),
         prevstrk = as.factor(prevstrk),
         prevhyp = as.factor(prevhyp),
         sysbp = as.numeric(levels(sysbp))[sysbp],
         diabp = as.numeric(as.character(diabp)),
         bmi = as.numeric(as.character(bmi))) %>% 
  as.tibble()
```

# Descriptive statistics
```{r}
#descriptive statistics for the data

summary(frmgham)
```



#missing values
```{r}
frmgham %>%
  group_by(period) %>%
  summarize(sum_na_period = sum(is.na(period)))

d1 = frmgham %>%
  group_by(randid) %>%
  summarize(n = n())
table(d1$n)
# 447 patients have only one obs
# 781 patients have only two obs
# 3206 patients have all three obs

number.subjects <-
  frmgham  %>%
  select(randid) %>%
  distinct() %>%
  unlist() %>%
  length()

number.subjects

# missing data pattern by variables

sort(sapply(frmgham, function(x) { sum(is.na(x)) }), decreasing=TRUE)

```



* The output tells us that 2243 samples are complete, 7077 samples miss both hdlc and ldlc, 4 samples miss only the glucose value and so on.


* there are 447 patients who have only one obs; there are 781 patients who have only two obs and 3206 patients have all three obs. In total, there are 4434 patients in the study.

##1. Target question:
1) Is there a relationship between age and smoking status?  Does this relationship differ by sex? 


### cursmoke vs age_ctg without adjustment
```{r}
## created age category variable (age_ctg)
frmgham =
  frmgham %>% 
  mutate(
      age_ctg = case_when(
      age < 35 ~ "35-",  
      age %in% 35:44 ~ "35-44",
      age %in% 45:54 ~ "45-54",
      age %in% 55:64 ~ "55-64",
      age %in% 65:74 ~ "65-74",
      age >= 75 ~ "75+"
    ),
    age_ctg = fct_relevel(age_ctg, "35-")) 

### cursmoke vs age_ctg : associated without adjustment 
frmgham %>% 
  select(randid, cursmoke, age_ctg) %>% 
  na.omit() %>% 
  group_by(age_ctg) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n()) %>% 
  ggplot(aes(x = age_ctg, y = p_cursmoke)) + 
  geom_bar(stat = "identity") +
  labs(title = "Relationship between categorized age and smoking status",
       x = "Age",
       y = "Probability of current smoking")
```

## Finding confounders 
### age and cursmoke -- confounder: hdlc, ldlc

```{r}
age <-
  glmer(cursmoke ~ age + (1|randid),
      data = frmgham,
      family = binomial, 
      nAGQ = 0)

pot_conf <- names(frmgham)[-c(1:3, 5)]
pot_conf <-
  pot_conf[-c(grep("time", pot_conf))]

conf_det <- vector("list", length = length(pot_conf))
names(conf_det) <- pot_conf



for (conf in pot_conf) {

  data <- cbind(frmgham[,c("cursmoke", "age", "randid")], frmgham[,conf])
  formula <- paste("cursmoke","~","age + (1|randid) +", conf)
  try(fit_glmer <- glmer(formula,
      data = data,
      family = binomial,
      nAGQ = 0))

  result <- tibble(
    variable = c("age"),
    OR =
      exp(coef(summary(age))[2,1]),
    lower_CI =
      exp(coef(summary(age))[2,1] - 1.96*(coef(summary(age))[2,2])),
    upper_CI =
      exp(coef(summary(age))[2,1] + 1.96*(coef(summary(age))[2,2])),
    potconf_OR =
      exp(coef(summary(fit_glmer))[2,1]),
    confounder =
      if_else((potconf_OR >= lower_CI & potconf_OR <= upper_CI), FALSE, TRUE)
  )

  conf_det[[conf]] <- result
}

map(conf_det, knitr::kable, digits = 3)
```




## Finding modifier

```{r}
### by sex -- modified!
frmgham %>% 
  mutate(sex = ifelse(sex == 1, "male", "women")) %>%
  select(randid, cursmoke, age_ctg, sex) %>% 
  na.omit() %>% 
  group_by(age_ctg, sex) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n(),
            lower_CI = prop.test(sum(cursmoke), n())$conf.int[1],
            upper_CI = prop.test(sum(cursmoke), n())$conf.int[2]) %>%
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = sex, group = sex)) +
  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI), width = .1) +
  geom_line() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")


### by educ -- not modified except 35-

frmgham %>% 
  select(randid, cursmoke, age_ctg, educ) %>% 
  na.omit() %>% 
  group_by(age_ctg, educ) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n()) %>% 
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = educ, group = educ)) + 
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")

### by diabetes -- not modified 

frmgham %>% 
  select(randid, cursmoke, age_ctg, diabetes) %>% 
  na.omit() %>% 
  group_by(age_ctg, diabetes) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n()) %>% 
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = diabetes, group = diabetes)) + 
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")

### by prevap -- not modified 

frmgham %>% 
  select(randid, cursmoke, age_ctg, prevap) %>% 
  na.omit() %>% 
  group_by(age_ctg, prevap) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n()) %>% 
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = prevap, group = prevap)) + 
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")

### by prevchd -- not modified 

frmgham %>% 
  select(randid, cursmoke, age_ctg, prevchd) %>% 
  na.omit() %>% 
  group_by(age_ctg, prevchd) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n()) %>% 
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = prevchd, group = prevchd)) + 
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")

### by prevmi -- not modified 

frmgham %>% 
  select(randid, cursmoke, age_ctg, prevmi) %>% 
  na.omit() %>% 
  group_by(age_ctg, prevmi) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n(),
            lower_CI = prop.test(sum(cursmoke), n())$conf.int[1],
            upper_CI = prop.test(sum(cursmoke), n())$conf.int[2]) %>%
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = prevmi, group = prevmi)) +
  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI), width = .1) +
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")

### by prevstrk -- not modified 

frmgham %>% 
  select(randid, cursmoke, age_ctg, prevstrk) %>% 
  na.omit() %>% 
  group_by(age_ctg, prevstrk) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n(),
            lower_CI = prop.test(sum(cursmoke), n())$conf.int[1],
            upper_CI = prop.test(sum(cursmoke), n())$conf.int[2]) %>%
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = prevstrk, group = prevstrk)) +
  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI), width = .1) +
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")

### by prevhyp -- slightly modified 

frmgham %>% 
  select(randid, cursmoke, age_ctg, prevhyp) %>% 
  na.omit() %>% 
  group_by(age_ctg, prevhyp) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n(),
            lower_CI = prop.test(sum(cursmoke), n())$conf.int[1],
            upper_CI = prop.test(sum(cursmoke), n())$conf.int[2]) %>%
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = prevhyp, group = prevhyp)) +
  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI), width = .1) + 
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")
```

## Fit model using glmer
```{r}
#educ variable
##https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4939617/
  
cursmoke_age <-
  glmer(cursmoke ~ age*sex + educ + hdlc + ldlc + age*prevhyp + (1|randid), 
        data = frmgham, family = binomial, nAGQ = 0)

summary(cursmoke_age) ## AIC 3581.6, var(random) = 0.007

# changed model based on cigpday
cursmoke_age <-
  glmer(cursmoke ~ age*sex + educ + heartrte + hdlc + ldlc + age*prevhyp + (1|randid), 
        data = frmgham, family = binomial, nAGQ = 0)

summary(cursmoke_age) ## AIC 3550.2, var(random) = 0.004

# cursmoke_age <-
#   glmer(cursmoke ~ age*sex + educ + heartrte + age*prevhyp + (1|randid), 
#         data = frmgham, family = binomial, nAGQ = 0)
# 
# summary(cursmoke_age) ## AIC 10644.9 var(random) = 21.15
```

```{r}
# cigpday
frmgham_smoker =
  frmgham %>% filter(cigpday != 0)

cigpday_age <-
  glmer(cigpday ~ age*sex + educ + heartrte + age*prevhyp + (1|randid), 
        data = frmgham_smoker, family = poisson, nAGQ = 0)

summary(cigpday_age)
```

(2) Is there a relationship between the number of cigarettes smoked per day and age?  Does this relationship differ by sex? 

While answering these questions, please account for any confounders that you have evidence may impact the relationship between age and sex with smoking.

```{r}
frmgham %>% 
  ggplot(., aes(x=age, y=cigpday)) +
    geom_jitter() + 
    geom_smooth(method = "loess") +
  ggtitle("Unadjusted relationship between the number of cigarettes smoked per day and age")
```

```{r}
frmgham_smoker  = frmgham %>%
filter(cigpday != 0) 

frmgham_smoker %>%
  pull(cigpday) %>%
  summary()

frmgham_smoker %>%
 group_by(randid) %>%
 mutate(mean.cigpday = mean(cigpday)) %>%
 ungroup() %>%
 select(randid, mean.cigpday) %>%
 distinct() %>%
 ggplot(aes(x = mean.cigpday)) +
 geom_histogram(color="darkblue", fill="lightblue", aes(y=..density..)) +
 labs(x = "Average number of cigarettes smoked per day",
      y = "Density",
      title = "cigpday summary")
  
```

```{r finding counfounders for m2}
# I will look into SYSBP DIABP TOTCHOL BMI DIABETES 
# based on https://tobaccocontrol.bmj.com/content/14/5/315

age <- 
  gee(cigpday ~ age, 
      data = frmgham, 
      id = randid, 
      family = poisson,
      corstr = ("unstructured"))

pot_conf <- names(frmgham)[-c(2, 3, 5, 8)]
pot_conf <-
  pot_conf[-c(grep("time", pot_conf))]

conf_det <- vector("list", length = length(pot_conf))
names(conf_det) <- pot_conf

conf_det <- vector("list", length = length(pot_conf))
names(conf_det) <- pot_conf

for (conf in pot_conf) {

  data <- cbind(frmgham[,c("cursmoke", "age")], frmgham[,conf])
  try(fit_gee <- gee(cursmoke ~ .,
      data = data,
      id = frmgham$randid,
      family = poisson,
      corstr = ("unstructured")))

  result <- tibble(
    variable = c("age"),
    OR =
      exp(coef(summary(age))[2,1]),
    lower_CI  =
      exp(coef(summary(age))[2,1] - 1.96*(coef(summary(age))[2,4])),
    upper_CI =
      exp(coef(summary(age))[2,1] + 1.96*(coef(summary(age))[2,4])),
    potconf_OR =
      exp(coef(summary(fit_gee))[2,1]),
    confounder =
      if_else((potconf_OR >= lower_CI & potconf_OR <= upper_CI), FALSE, TRUE)
  )

  conf_det[[conf]] <- result
}

map(conf_det, knitr::kable, digits = 3)
map(conf_det, knitr::kable, digits = 3)
```

```{r backward selection}
cigpday.lmer1<- glmer(cigpday ~ age + sex + heartrte + hdlc + ldlc + totchol + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer1)
# AIC   7876.2 
cigpday.lmer2<- glmer(cigpday ~ age + sex + heartrte + hdlc  + totchol + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer2)
# AIC 7874.4
cigpday.lmer3<- glmer(cigpday ~ age + sex + heartrte + hdlc + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer3)
# AIC 7872.8

# tested interaction term with age
cigpday.lmer4<- glmer(cigpday ~ age + sex + heartrte + hdlc + age:heartrte + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer4)
# decided with heartrte
# AIC   7870.3
```


```{r final model }
cigpday.lmer<- glmer(cigpday ~ age + sex + heartrte + hdlc + age*heartrte + (1 | randid), data = frmgham_smoker, family = poisson, nAGQ = 0)
summary(cigpday.lmer)
```


#2.Target question:

Next you are interested in the relationship between certain health outcomes and smoking status. In particular you are interested in :


(1)The relationship between current smoking status and systolic blood pressure.
```{r}
frmgham %>% 
  ggplot(., aes(x=sysbp, y=cursmoke)) +
    geom_jitter() + 
    geom_smooth(method = "loess") + 
  ggtitle("Unadjusted relationship between current smoking status and diastolic blood pressure")

#https://www.ncbi.nlm.nih.gov/pubmed/7790985
#From the research paper above, diastolic blood pressure and hypertension are potential confounders.

#center the variables and find the confounder
frmgham$diabp_c = frmgham$diabp-mean(frmgham$diabp)
frmgham$sysbp_c = frmgham$sysbp-mean(frmgham$sysbp)

sysbp <- 
  glmer(cursmoke ~ sysbp_c + (1|randid), 
      data = frmgham, 
      family = binomial)

sysbp_diabp <- 
  glmer(cursmoke ~ sysbp_c + diabp_c + (1|randid), 
      data = frmgham, 
      family = binomial)

sysbp_prevhyp <- 
  glmer(cursmoke ~ sysbp_c + prevhyp + (1|randid), 
      data = frmgham, 
      family = binomial)

sysbp_hyperten <- 
  glmer(cursmoke ~ sysbp_c + hyperten+ (1|randid), 
      data = frmgham, 
      family = binomial)

tibble(
    variable = c("diabp", "prevhyp", "hyperten"),
    OR = 
      exp(coef(summary(sysbp))[2,1]),
    lower_CI  = 
      exp(coef(summary(sysbp))[2,1] - 1.96*(coef(summary(sysbp))[2,4])),
    upper_CI = 
      exp(coef(summary(sysbp))[2,1] + 1.96*(coef(summary(sysbp))[2,4])),
    conf_OR =
      c(exp(coef(summary(sysbp_diabp))[2,1]),
        exp(coef(summary(sysbp_prevhyp))[2,1]),
        exp(coef(summary(sysbp_hyperten))[2,1])),
    confounder =
      if_else((conf_OR >= lower_CI & conf_OR <= upper_CI), FALSE, TRUE)
  )

smoke_diabp = glmer(cursmoke ~ diabp_c + sysbp_c + prevhyp + hyperten+ (1|randid), 
      data = frmgham, 
      family = binomial)
summary(smoke_diabp)
```

(2)The relationship between current smoking status and diastolic blood pressure.

```{r}
#jitter plot of cursmoke and sysbp
frmgham %>% 
  ggplot(., aes(x=cursmoke, y=sysbp)) +
  geom_jitter() + 
  geom_smooth(method = "loess") + 
  ggtitle("Unadjusted relationship between current smoking status and systolic blood pressure")

frmgham = frmgham %>% mutate(cursmoke = as.factor(cursmoke),
                             period = as.factor(period))

##check random effects
ggplot(data = frmgham, aes(x = period, y = sysbp, group = randid)) +
  geom_point() +
  geom_line() +
  facet_wrap( ~ cursmoke, ncol=2) +
  ggtitle("sysbp versus period for smoker and nonsmoker")

syslm = lm(sysbp~cursmoke, data = frmgham)
sysbp <- lmer(sysbp ~ cursmoke + (1|randid), data = frmgham)
# exactLRT(sysbp, syslm)
##p-value < 2.2e-16, random effect is significant.

#find the confounder
sysbp_sex <- lmer(sysbp ~ cursmoke + sex + (1|randid), data = frmgham)
sysbp_totchol <- lmer(sysbp ~ cursmoke + totchol + (1|randid), data = frmgham)
sysbp_age <- lmer(sysbp ~ cursmoke + age + (1|randid), data = frmgham)
sysbp_diabp <- lmer(sysbp ~ cursmoke + diabp + (1|randid), data = frmgham)
sysbp_bmi <- lmer(sysbp ~ cursmoke + bmi + (1|randid), data = frmgham)
sysbp_diab <- lmer(sysbp ~ cursmoke + diabetes + (1|randid), data = frmgham)
sysbp_bpmeds <- lmer(sysbp ~ cursmoke + bpmeds + (1|randid), data = frmgham)
sysbp_heart <- lmer(sysbp ~ cursmoke + heartrte + (1|randid), data = frmgham)
sysbp_glu <- lmer(sysbp ~ cursmoke + glucose + (1|randid), data = frmgham)
sysbp_educ <- lmer(sysbp ~ cursmoke + educ + (1|randid), data = frmgham)
sysbp_prevchd <- lmer(sysbp ~ cursmoke + prevchd + (1|randid), data = frmgham)
sysbp_prevap <- lmer(sysbp ~ cursmoke + prevap + (1|randid), data = frmgham)
sysbp_prevmi <- lmer(sysbp ~ cursmoke + prevmi + (1|randid), data = frmgham)
sysbp_strk<- lmer(sysbp ~ cursmoke + prevstrk + (1|randid), data = frmgham)
sysbp_hyp <- lmer(sysbp ~ cursmoke + prevhyp + (1|randid), data = frmgham)
sysbp_per <- lmer(sysbp ~ cursmoke + period + (1|randid), data = frmgham)
tibble(
  variable = c("sex", "totchol", "age", "diabp", "bmi", "diab", "bpmeds",
               "heart", "glu", "educ", "prevchd", "prevap", "prevmi", "prevstrk", "prevhyp",
               "period"),
  OR = 
    exp(coef(summary(sysbp))[2,1]),
  lower_CI  = 
    exp(coef(summary(sysbp))[2,1] - 1.96*(coef(summary(sysbp))[2,2])),
  upper_CI = 
    exp(coef(summary(sysbp))[2,1] + 1.96*(coef(summary(sysbp))[2,2])),
  conf_OR =
    c(exp(coef(summary(sysbp_sex))[2,1]),
      exp(coef(summary(sysbp_totchol))[2,1]),
      exp(coef(summary(sysbp_age))[2,1]),
      exp(coef(summary(sysbp_diabp))[2,1]),
      exp(coef(summary(sysbp_bmi))[2,1]),
      exp(coef(summary(sysbp_diab))[2,1]),
      exp(coef(summary(sysbp_bpmeds))[2,1]),
      exp(coef(summary(sysbp_heart))[2,1]),
      exp(coef(summary(sysbp_glu))[2,1]),
      exp(coef(summary(sysbp_educ))[2,1]),
      exp(coef(summary(sysbp_prevchd))[2,1]),
      exp(coef(summary(sysbp_prevap))[2,1]),
      exp(coef(summary(sysbp_prevmi))[2,1]),
      exp(coef(summary(sysbp_strk))[2,1]),
      exp(coef(summary(sysbp_hyp))[2,1]),
      exp(coef(summary(sysbp_per))[2,1])),
  confounder =
    if_else((conf_OR >= lower_CI & conf_OR <= upper_CI), FALSE, TRUE)
)

#model adding the confounders
sysbp_smoke = lmer(sysbp ~ cursmoke+ age + diabp + bmi + prevhyp+ period + (1|randid), 
                    data = frmgham)

#check interaction
sysbp1 = lmer(sysbp ~ cursmoke*age + diabp + bmi + prevhyp+ period + (1|randid), 
              data = frmgham)
sysbp2 = lmer(sysbp ~ cursmoke*diabp + age + bmi + prevhyp +period+ (1|randid), 
              data = frmgham)
sysbp3 = lmer(sysbp ~ cursmoke*bmi + age + diabp + prevhyp+period + (1|randid), 
              data = frmgham)
sysbp4 = lmer(sysbp ~ cursmoke*prevhyp + age + bmi + diabp +period+ (1|randid), 
              data = frmgham)
sysbp5 = lmer(sysbp ~ cursmoke*period + age + bmi + diabp +prevhyp+ (1|randid), 
              data = frmgham)
summary(sysbp1)
summary(sysbp2)
summary(sysbp3)
summary(sysbp4)
summary(sysbp5)
AIC(sysbp1)
#interaction term reference: Ulrich John, Monika Hanke,Christian Meyer,Anja Schumann. 
#Gender and age differences among current smokers in a general population survey. BMC Public Health. 2005; 5: 57.
#Published online 2005 Jun 3. doi:  [10.1186/1471-2458-5-57]
AIC(sysbp2)
AIC(sysbp3)
AIC(sysbp4)
AIC(sysbp5)

#95% CI for beta of cursmoke and interaction term
b_par<-bootMer(x=sysbp1,FUN=fixef,nsim=200)
boot::boot.ci(b_par,type="basic",index=2)
b_par<-bootMer(x=sysbp1,FUN=fixef,nsim=200)
boot::boot.ci(b_par,type="basic",index=9)
```




##########################################################################################
#(2)The relationship between current smoking status and diastolic blood pressure.
```{r}
#jitter plot of cursmoke and diabp
frmgham %>% 
  ggplot(., aes(x=cursmoke, y=diabp)) +
  geom_jitter() + 
  geom_smooth(method = "loess") + 
  ggtitle("Unadjusted relationship between current smoking status and diastolic blood pressure")

#check random effects
ggplot(data = frmgham, aes(x = period, y = diabp, group = randid)) +
  geom_point() +
  geom_line() +
  facet_wrap( ~ cursmoke, ncol=2) +
  ggtitle("diabp versus period for smoker and nonsmoker")

dialm = lm(diabp~cursmoke, data = frmgham)
diabp <- lmer(diabp ~ cursmoke + (1|randid), data = frmgham)
# exactLRT(diabp, dialm)
#p-value < 2.2e-16, random effect is significant.

#find confounder
diabp_sex <- lmer(diabp ~ cursmoke + sex + (1|randid), data = frmgham)
diabp_totchol <- lmer(diabp ~ cursmoke + totchol + (1|randid), data = frmgham)
diabp_age <- lmer(diabp ~ cursmoke + age + (1|randid), data = frmgham)
diabp_sysbp <- lmer(diabp ~ cursmoke + sysbp + (1|randid), data = frmgham)
diabp_bmi <- lmer(diabp ~ cursmoke + bmi + (1|randid), data = frmgham)
diabp_diab <- lmer(diabp ~ cursmoke + diabetes + (1|randid), data = frmgham)
diabp_bpmeds <- lmer(diabp ~ cursmoke + bpmeds + (1|randid), data = frmgham)
diabp_heart <- lmer(diabp ~ cursmoke + heartrte + (1|randid), data = frmgham)
diabp_glu <- lmer(diabp ~ cursmoke + glucose + (1|randid), data = frmgham)
diabp_educ <- lmer(diabp ~ cursmoke + educ + (1|randid), data = frmgham)
diabp_prevchd <- lmer(diabp ~ cursmoke + prevchd + (1|randid), data = frmgham)
diabp_prevap <- lmer(diabp ~ cursmoke + prevap + (1|randid), data = frmgham)
diabp_prevmi <- lmer(diabp ~ cursmoke + prevmi + (1|randid), data = frmgham)
diabp_strk<- lmer(diabp ~ cursmoke + prevstrk + (1|randid), data = frmgham)
diabp_hyp <- lmer(diabp ~ cursmoke + prevhyp + (1|randid), data = frmgham)
diabp_per <- lmer(diabp ~ cursmoke + period + (1|randid), data = frmgham)
tibble(
  variable = c("sex", "totchol", "age", "sysbp", "bmi", "diab", "bpmeds",
               "heart", "glu", "educ", "prevchd", "prevap", "prevmi", "prevstrk", "prevhyp",
               "period"),
  OR = 
    exp(coef(summary(diabp))[2,1]),
  lower_CI  = 
    exp(coef(summary(diabp))[2,1] - 1.96*(coef(summary(diabp))[2,2])),
  upper_CI = 
    exp(coef(summary(diabp))[2,1] + 1.96*(coef(summary(diabp))[2,2])),
  conf_OR =
    c(exp(coef(summary(diabp_sex))[2,1]),
      exp(coef(summary(diabp_totchol))[2,1]),
      exp(coef(summary(diabp_age))[2,1]),
      exp(coef(summary(diabp_sysbp))[2,1]),
      exp(coef(summary(diabp_bmi))[2,1]),
      exp(coef(summary(diabp_diab))[2,1]),
      exp(coef(summary(diabp_bpmeds))[2,1]),
      exp(coef(summary(diabp_heart))[2,1]),
      exp(coef(summary(diabp_glu))[2,1]),
      exp(coef(summary(diabp_educ))[2,1]),
      exp(coef(summary(diabp_prevchd))[2,1]),
      exp(coef(summary(diabp_prevap))[2,1]),
      exp(coef(summary(diabp_prevmi))[2,1]),
      exp(coef(summary(diabp_strk))[2,1]),
      exp(coef(summary(diabp_hyp))[2,1]),
      exp(coef(summary(diabp_per))[2,1])),
  confounder =
    if_else((conf_OR >= lower_CI & conf_OR <= upper_CI), FALSE, TRUE)
)

#model adding the confounders
diabp_smoke = lmer(diabp ~ cursmoke + sysbp + bmi + prevhyp + (1|randid), 
                   data = frmgham)
summary(diabp_smoke)

#check interaction
diabp1 = lmer(diabp ~ cursmoke*sysbp + bmi + prevhyp + (1|randid), 
              data = frmgham)
diabp2 = lmer(diabp ~ cursmoke*bmi +sysbp+ prevhyp + (1|randid), 
              data = frmgham)
diabp3 = lmer(diabp ~ cursmoke*prevhyp + bmi + sysbp +(1|randid), 
              data = frmgham)
summary(diabp1)
summary(diabp2)
summary(diabp3)
AIC(diabp1)
AIC(diabp2)
AIC(diabp3)
AIC(diabp_smoke)

#95% CI of beta of cursmoke and interaction term
b_par<-bootMer(x=diabp1,FUN=fixef,nsim=200)
boot::boot.ci(b_par,type="basic",index=2)
b_par<-bootMer(x=diabp1,FUN=fixef,nsim=200)
boot::boot.ci(b_par,type="basic",index=6)
```
From the table above, we found "sysbp", "prevhyp" and "hyperten" are confounders. After adding the confounders, the final model is shown above. For a one-unit increase in the diastolic blood pressure, the expected change in log odds is 0.009, adjusting for systolic blood pressure, prevalent hypertensive and hypertensive.


---

(3)The relationship between current smoking status and serum total cholesterol. Again, while answering these questions, please account for any confounders that you have evidence may impact these relationships. 


```{r}
frmgham %>% 
  ggplot(., aes(x=cursmoke, y=totchol)) +
    geom_bar(stat="identity") +ggtitle("Unadjusted relationship between current smoking status and serum total cholesterol")
frmgham %>% 
  ggplot(., aes(x=as.factor(cursmoke), y=totchol)) +
  geom_boxplot() +ggtitle("Serum total cholesterol by current smoking status, Boxplot")
frmgham %>% 
  ggplot(., aes(x=as.factor(cursmoke), y=totchol)) +
  geom_violin(aes(fill = as.factor(cursmoke))) +ggtitle("Serum total cholesterol by current smoking status, Violin")
#https://medlineplus.gov/cholesterollevelswhatyouneedtoknow.html
#categorized totchol based on link above
frmgham=frmgham %>% 
  mutate(totchol_ctg= ifelse(age<=19 & totchol<170,0,
                             ifelse(age>19&totchol %in% c(125, 200), 0,1)))
table(frmgham$totchol_ctg)
# 124 normal total chol 
# 11503 abnormal total chol
```
## Finding confounders
```{r}
library(nlme)
frmgham= frmgham %>% 
  mutate(bmi=as.numeric(as.character(bmi))) %>% 
  mutate(bmi_cat = ifelse(bmi<18.5,0,ifelse(bmi %in% c(18.5,24.9), 1, ifelse(bmi %in% c(25, 29.9),2, 3))))
a=as.data.frame(lapply(frmgham[,c("cursmoke","age","sex","prevchd")], as.numeric))
cor(a)
totchol1 <- 
  gls(totchol ~ as.factor(cursmoke)*period+sex+bmi+age, 
      data = frmgham,na.action=na.omit, 
      corr = corCompSymm(,form= ~ 1 | randid) )
totchol <- 
  gls(totchol ~ as.factor(cursmoke), data = frmgham,na.action=na.omit)
totchol3=lmer(totchol ~ as.factor(cursmoke)+age+sex+prevchd+(1|randid),
              data=frmgham)
totchol4=lmer(totchol ~ as.factor(cursmoke)*age+sex+prevchd+(as.factor(cursmoke)|randid),
              data=frmgham)
totchol5=lmer(totchol ~ as.factor(cursmoke)+age+sex+prevchd+bmi+(1|randid),
              data=frmgham)
totchol6=lmer(totchol ~ as.factor(cursmoke)+age+sex+prevchd+bmi+(1|randid),
              data=frmgham)
frmgham=frmgham %>% 
  mutate(period = period-1)
totchol7=lmer(totchol ~ as.factor(cursmoke)*sex+bmi+age+prevchd+(1|randid),
              data=frmgham)
totchol8=lmer(totchol ~ as.factor(cursmoke)*sex+bmi+age+(1|randid),
              data=frmgham)
totchol9=lmer(totchol ~ as.factor(cursmoke)+sex+bmi+age+(1|randid),
              data=frmgham)
totchol11=lmer(totchol ~ as.factor(cursmoke)*as.factor(period)+sex+bmi+age+(1|randid),
              data=frmgham)
totchol10=lmer(totchol ~ as.factor(cursmoke)+as.factor(period)+sex+bmi+age+(1|randid),
              data=frmgham)
totchol12=lmer(totchol ~ as.factor(cursmoke)+as.factor(period)+bmi+age+(1|randid),
              data=frmgham)
#final model
totchol13=lmer(totchol ~ as.factor(cursmoke)*sex+bmi+age+prevstrk+prevchd+(1|randid),
              data=frmgham)
summary(totchol13)
# AIC(totchol10)
# summary(totchol10)
# summary(totchol11)
# anova(totchol13,totchol14)
# test the significance of random effect
#sig_ranef=RLRsim::exactRLRT(totchol13, seed=1)
# pot_conf <- names(frmgham)[-c(1,2,4)]
# pot_conf <- pot_conf[-c(grep("time", pot_conf), grep("death", pot_conf))]
# 
# conf_det <- vector("list", length = length(pot_conf))
# names(conf_det) <- pot_conf
# 
# 
# 
# for (conf in pot_conf) {
# 
#  data <- cbind(frmgham[,c("totchol","cursmoke")], frmgham[,conf])
#  try(fit_gee <- gls(totchol ~ .,
#      data = data,
#      ,na.action=na.omit))
# 
#  result <- tibble(
#    variable = c("totchol"),
#    OR =
#      exp(coef(summary(totchol))[2,1]),
#    lower_CI  =
#      exp(coef(summary(totchol))[2,1] - 1.96*(coef(summary(totchol))[2,3])),
#    upper_CI =
#      exp(coef(summary(totchol))[2,1] + 1.96*(coef(summary(totchol))[2,3])),
#    potconf_OR =
#      exp(coef(summary(fit_gee))[2,1]),
#    confounder =
#      if_else((potconf_OR >= lower_CI & potconf_OR <= upper_CI), FALSE, TRUE)
#  )
# 
#  conf_det[[conf]] <- result
# }
# 
# map(conf_det, knitr::kable, digits = 3)
```
```{r}
plot_dt=frmgham %>%
  mutate(cursmoke=as.factor(cursmoke)) %>% 
  group_by(cursmoke, sex) %>%
  summarise(
    n=n(),
    mean=mean(totchol, na.rm=T),
    sd=sd(totchol, na.rm=T)
  ) %>%
  mutate( se=sd/sqrt(n))  %>%
  mutate( ic=se * qt((1-0.05)/2 + .5, n-1))
ggplot(plot_dt)+
  geom_bar( aes(x=cursmoke, y=mean), stat="identity", fill="forestgreen", alpha=0.5) +
  geom_errorbar( aes(x=cursmoke, ymin=mean-ic, ymax=mean+ic), width=0.4, colour="orange", alpha=0.9, size=1.5) +
  ggtitle("Mean totochol by gender")+facet_wrap(~sex)

frmgham %>%
  mutate(cursmoke=as.factor(cursmoke)) %>% 
  group_by(cursmoke, prevchd) %>%
  summarise(
    n=n(),
    mean=mean(totchol, na.rm=T),
    sd=sd(totchol, na.rm=T)
  ) %>%
  mutate( se=sd/sqrt(n))  %>%
  mutate( ic=se * qt((1-0.05)/2 + .5, n-1)) %>% 
  ggplot(.)+
  geom_bar(aes(x=cursmoke, y=mean), stat="identity", fill="forestgreen", alpha=0.5) +
  geom_errorbar( aes(x=cursmoke, ymin=mean-ic, ymax=mean+ic), width=0.4, colour="orange", alpha=0.9, size=1.5) +
  ggtitle("Mean totochol by prevalence of coronary heart disease ")+facet_wrap(~prevchd)
frmgham %>%
  mutate(cursmoke=as.factor(cursmoke)) %>% 
  group_by(cursmoke, age_ctg) %>%
  summarise(
    n=n(),
    mean=mean(totchol, na.rm=T),
    sd=sd(totchol, na.rm=T)
  ) %>%
  mutate( se=sd/sqrt(n))  %>%
  mutate( ic=se * qt((1-0.05)/2 + .5, n-1)) %>% 
  ggplot(.)+
  geom_bar(aes(x=cursmoke, y=mean), stat="identity", fill="forestgreen", alpha=0.5) +
  geom_errorbar( aes(x=cursmoke, ymin=mean-ic, ymax=mean+ic), width=0.4, colour="orange", alpha=0.9, size=1.5) +
  ggtitle("Mean totochol by age category")+facet_wrap(~age_ctg)
frmgham %>%
  mutate(cursmoke=as.factor(cursmoke)) %>% 
  group_by(cursmoke, bmi_cat) %>%
  summarise(
    n=n(),
    mean=mean(totchol, na.rm=T),
    sd=sd(totchol, na.rm=T)
  ) %>%
  mutate( se=sd/sqrt(n))  %>%
  mutate( ic=se * qt((1-0.05)/2 + .5, n-1)) %>% 
  ggplot(.)+
  geom_bar(aes(x=cursmoke, y=mean), stat="identity", fill="forestgreen", alpha=0.5) +
  geom_errorbar( aes(x=cursmoke, ymin=mean-ic, ymax=mean+ic), width=0.4, colour="orange", alpha=0.9, size=1.5) +
  ggtitle("Mean totochol by BMI category")+facet_wrap(~bmi_cat)
```
From the table above, we found age, ldlc and hdlc are confounders. 
## bootstrap for confidence interval
```{r}
b_par<-bootMer(x=totchol13,FUN=fixef,nsim=200)
boot::boot.ci(b_par,type="basic",index=2)
```



```{r}
lmer <- lmer(diabp ~ cursmoke*sysbp + bmi + prevhyp + (1|randid),  data = frmgham)
summary(lmer)
```







