---
title: "logitudinal_final_proj"
author: "Lynette Pan"
date: "12/3/2018"
output: pdf_document
---

#Final Project: Smoking in the Framingham Heart Study


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(janitor)
library(mice)
library(gee)

frmgham = 
  read.csv2("frmgham2.csv", sep = ",", na.strings=c("","NA")) %>% 
  clean_names() %>% select(cursmoke, everything()) %>% 
  mutate(sex = as.factor(sex),
         bpmeds = as.factor(bpmeds),
         educ = as.factor(educ),
         diabetes = as.factor(diabetes),
         prevap = as.factor(prevap),
         prevchd = as.factor(prevchd),
         prevmi = as.factor(prevmi),
         prevstrk = as.factor(prevstrk),
         prevhyp = as.factor(prevhyp),
         sysbp = as.numeric(levels(sysbp))[sysbp],
         diabp = as.numeric(as.character(diabp)),
         bmi = as.numeric(as.character(bmi))) %>% 
  as.tibble()

## created age category variable (age_ctg)
frmgham =
  frmgham %>% 
  mutate(
      age_ctg = case_when(
      age < 35 ~ "35-",  
      age %in% 35:44 ~ "35-44",
      age %in% 45:54 ~ "45-54",
      age %in% 55:64 ~ "55-64",
      age %in% 65:74 ~ "65-74",
      age >= 75 ~ "75+"
    ),
    age_ctg = fct_relevel(age_ctg, "35-")) 

## created sysbp category variable (sysbp_ctg)
### According to the new guideline from american heart association
### https://www.ahajournals.org/doi/10.1161/HYP.0000000000000065
frmgham =
  frmgham %>% 
  mutate(
      sysbp_ctg = case_when(
      sysbp < 120 ~ "normal",  
      (sysbp >= 120 & sysbp < 130) ~ "elevated",
      (sysbp >= 130 & sysbp < 140) ~ "stage1",
      (sysbp >= 140 & sysbp <= 180) ~ "stage2",
      sysbp > 180 ~ "crisis"
    ),
    sysbp_ctg = fct_relevel(sysbp_ctg, "normal")) 

## created sysbp category variable (diabp_ctg)
### According to the new guideline from american heart association
### https://www.ahajournals.org/doi/10.1161/HYP.0000000000000065
frmgham =
  frmgham %>% 
  mutate(
      diabp_ctg = case_when(
      diabp < 80 ~ "normal and elevated",  
      (diabp >= 80 & diabp < 90) ~ "stage1",
      (diabp >= 90 & diabp <= 120 ) ~ "stage2",
      diabp > 120 ~ "crisis"
      ),
    diabp_ctg = fct_relevel(diabp_ctg, "normal and elevated")) 

```

# Descriptive statistics
```{r}
#descriptive statistics for the data

summary(frmgham)

frmgham %>% distinct(randid) %>% count()
```



#missing values
```{r}
# frmgham %>% 
#   group_by(period) %>% 
#   summarize(sum_na_period = sum(is.na(period)))
# 
# d1 = frmgham %>% 
#   group_by(randid) %>% 
#   summarize(n = n())
# table(d1$n)
# # 447 patients have only one obs
# # 781 patients have only two obs
# # 3206 patients have all three obs
# 
# number.subjects <- 
#   frmgham  %>% 
#   select(randid) %>% 
#   distinct() %>% 
#   unlist() %>%
#   length()
# 
# number.subjects
# 
# # missing data pattern by variables
# 
# sort(sapply(frmgham, function(x) { sum(is.na(x)) }), decreasing=TRUE)

```

# Impute missing values using MICE package
```{r}
load("frmgham_complete.RData")
# # frmgham_temp <- mice(frmgham,m=1, method='cart', printFlag=FALSE,seed = 1)
# # frmgham
# # summary(frmgham_temp)
# # frmgham_complete <- complete(frmgham_temp)
# #save(frmgham_complete, file = "frmgham_complete.RData")
# frmgham_complete =
#   frmgham_complete %>% 
#   mutate(
#       age_ctg = case_when(
#       age < 35 ~ "35-",  
#       age %in% 35:44 ~ "35-44",
#       age %in% 45:54 ~ "45-54",
#       age %in% 55:64 ~ "55-64",
#       age %in% 65:74 ~ "65-74",
#       age >= 75 ~ "75+"
#     )) 
# 
# sort(sapply(frmgham_complete, function(x) { sum(is.na(x)) }), decreasing=TRUE)

```


* The output tells us that 2243 samples are complete, 7077 samples miss both hdlc and ldlc, 4 samples miss only the glucose value and so on.


* there are 447 patients who have only one obs; there are 781 patients who have only two obs and 3206 patients have all three obs. In total, there are 4434 patients in the study.

##1. Target question:
1) Is there a relationship between age and smoking status?  Does this relationship differ by sex? 







(2) Is there a relationship between the number of cigarettes smoked per day and age?  Does this relationship differ by sex? 


## cigpday summary
```{r}
frmgham_smoker %>% 
  group_by(randid) %>% 
  mutate(mean.cigpday = mean(cigpday)) %>% 
  ungroup() %>% 
  select(randid, mean.cigpday) %>% 
  distinct() %>% 
  ggplot(aes(x = mean.cigpday)) + 
  geom_histogram() +
  labs(x = "Average number of cigarrette smoked each day",
       y = "Count",
       title = "cigpday summary")
```




```{r}
# I will look into SYSBP DIABP TOTCHOL BMI DIABETES 
# based on https://tobaccocontrol.bmj.com/content/14/5/315
```



```{r}
frmgham %>% 
  mutate(sex=ifelse(sex==1, "male","women")) %>% 
  ggplot(., aes(x=age, y=cigpday)) +
  geom_jitter() + 
  geom_smooth(method = "loess") +
  facet_grid(~sex)+
  ggtitle("Relationship between the number of cigarettes smoked per day and age by gender")
```




#2.Target question:

Next you are interested in the relationship between certain health outcomes and smoking status. In particular you are interested in :


(1)The relationship between current smoking status and systolic blood pressure.
```{r}
frmgham %>% 
  ggplot(., aes(x=sysbp, y=cursmoke)) +
    geom_jitter() + 
    geom_smooth(method = "loess") + 
  ggtitle("Unadjusted relationship between current smoking status and diastolic blood pressure")

#https://www.ncbi.nlm.nih.gov/pubmed/7790985
#From the research paper above, diastolic blood pressure and hypertension are potential confounders.

#center the variables and find the confounder
frmgham$diabp_c = frmgham$diabp-mean(frmgham$diabp)
frmgham$sysbp_c = frmgham$sysbp-mean(frmgham$sysbp)

sysbp <- 
  glmer(cursmoke ~ sysbp_c + (1|randid), 
      data = frmgham, 
      family = binomial)

sysbp_diabp <- 
  glmer(cursmoke ~ sysbp_c + diabp_c + (1|randid), 
      data = frmgham, 
      family = binomial)

sysbp_prevhyp <- 
  glmer(cursmoke ~ sysbp_c + prevhyp + (1|randid), 
      data = frmgham, 
      family = binomial)

sysbp_hyperten <- 
  glmer(cursmoke ~ sysbp_c + hyperten+ (1|randid), 
      data = frmgham, 
      family = binomial)

tibble(
    variable = c("diabp", "prevhyp", "hyperten"),
    OR = 
      exp(coef(summary(sysbp))[2,1]),
    lower_CI  = 
      exp(coef(summary(sysbp))[2,1] - 1.96*(coef(summary(sysbp))[2,4])),
    upper_CI = 
      exp(coef(summary(sysbp))[2,1] + 1.96*(coef(summary(sysbp))[2,4])),
    conf_OR =
      c(exp(coef(summary(sysbp_diabp))[2,1]),
        exp(coef(summary(sysbp_prevhyp))[2,1]),
        exp(coef(summary(sysbp_hyperten))[2,1])),
    confounder =
      if_else((conf_OR >= lower_CI & conf_OR <= upper_CI), FALSE, TRUE)
  )

smoke_diabp = glmer(cursmoke ~ diabp_c + sysbp_c + prevhyp + hyperten+ (1|randid), 
      data = frmgham, 
      family = binomial)
summary(smoke_diabp)
```

(2)The relationship between current smoking status and diastolic blood pressure.
```{r}
diabp <- 
  gee(cursmoke ~ diabp, 
      data = frmgham, 
      id = randid, 
      family = binomial,
      corstr = ("unstructured"))

diabp_sysbp <- 
  gee(cursmoke ~ diabp + sysbp, 
      data = frmgham, 
      id = randid, 
      family = binomial,
      corstr = ("unstructured"))

diabp_prevhyp <- 
  gee(cursmoke ~ diabp + prevhyp, 
      data = frmgham, 
      id = randid, 
      family = binomial,
      corstr = ("unstructured"))

diabp_hyperten <- 
  gee(cursmoke ~ diabp + hyperten, 
      data = frmgham, 
      id = randid, 
      family = binomial,
      corstr = ("unstructured"))

tibble(
    variable = c("sysbp", "prevhyp", "hyperten"),
    OR = 
      exp(coef(summary(diabp))[2,1]),
    lower_CI  = 
      exp(coef(summary(diabp))[2,1] - 1.96*(coef(summary(diabp))[2,4])),
    upper_CI = 
      exp(coef(summary(diabp))[2,1] + 1.96*(coef(summary(diabp))[2,4])),
    conf_OR =
      c(exp(coef(summary(diabp_sysbp))[2,1]),
        exp(coef(summary(diabp_prevhyp))[2,1]),
        exp(coef(summary(diabp_hyperten))[2,1])),
    confounder =
      if_else((conf_OR >= lower_CI & conf_OR <= upper_CI), FALSE, TRUE)
  )

smoke_diabp = gee(cursmoke ~ diabp + sysbp + prevhyp + hyperten, 
      data = frmgham, 
      id = randid, 
      family = binomial,
      corstr = ("unstructured"))
summary(smoke_diabp)

#confounders: sysbp, prevhyp, hyperten
```

From the table above, we found "sysbp", "prevhyp" and "hyperten" are confounders. After adding the confounders, the final model is shown above. For a one-unit increase in the diastolic blood pressure, the expected change in log odds is 0.009, adjusting for systolic blood pressure, prevalent hypertensive and hypertensive.


---

(3)The relationship between current smoking status and serum total cholesterol. Again, while answering these questions, please account for any confounders that you have evidence may impact these relationships. 


```{r}
frmgham %>% 
  ggplot(., aes(x=totchol, y=cursmoke)) +
    geom_jitter() + 
    geom_smooth(method = "loess") +ggtitle("Unadjusted relationship between current smoking status and serum total cholesterol")


#https://medlineplus.gov/cholesterollevelswhatyouneedtoknow.html
#categorized totchol based on link above

frmgham=frmgham %>% 
  mutate(totchol_ctg= ifelse(age<=19 & totchol<170,0,
                             ifelse(age>19&totchol %in% c(125, 200), 0,1)))
table(frmgham$totchol_ctg)
# 124 normal total chol 
# 11503 abnormal total chol
```

## Finding confounders


```{r}

frmgham= frmgham %>% 
  mutate(bmi=as.numeric(as.character(bmi))) %>% 
  mutate(bmi_cat = ifelse(bmi<18.5,0,ifelse(bmi %in% c(18.5,24.9), 1, ifelse(bmi %in% c(25, 29.9),2, 3))))

totchol <- 
  gee(cursmoke ~ totchol, 
      data = frmgham, 
      id = randid, 
      family = binomial,
      corstr = ("unstructured"))
pot_conf <- names(frmgham)[-c(1,2,4)]
pot_conf <- pot_conf[-c(grep("time", pot_conf), grep("death", pot_conf))]

conf_det <- vector("list", length = length(pot_conf))
names(conf_det) <- pot_conf



for (conf in pot_conf) {

 data <- cbind(frmgham[,c("cursmoke", "totchol")], frmgham[,conf])
 try(fit_gee <- gee(cursmoke ~ .,
     data = data,
     id = frmgham$randid,
     family = binomial,
     corstr = ("unstructured")))

 result <- tibble(
   variable = c("totchol"),
   OR =
     exp(coef(summary(totchol))[2,1]),
   lower_CI  =
     exp(coef(summary(totchol))[2,1] - 1.96*(coef(summary(totchol))[2,4])),
   upper_CI =
     exp(coef(summary(totchol))[2,1] + 1.96*(coef(summary(totchol))[2,4])),
   potconf_OR =
     exp(coef(summary(fit_gee))[2,1]),
   confounder =
     if_else((potconf_OR >= lower_CI & potconf_OR <= upper_CI), FALSE, TRUE)
 )

 conf_det[[conf]] <- result
}

map(conf_det, knitr::kable, digits = 3)
```


(1) age


```{r}


frmgham%>% 
  select(randid,totchol_ctg, age_ctg, cursmoke) %>% 
  na.omit() %>% 
  group_by(totchol_ctg,age_ctg) %>% 
  dplyr::summarise(percent.currentsmoke = sum(cursmoke)/n(),
                   n = n())
frmgham %>%
  select(randid,totchol_ctg, age_ctg, cursmoke) %>% 
  na.omit() %>% 
  group_by(totchol_ctg,age_ctg) %>%
  summarise(num.currentsmoke = sum(cursmoke)) %>% 
  ggplot(., aes(as.factor(totchol_ctg),num.currentsmoke)) +
    geom_bar(stat = "identity")+
  facet_wrap(~age_ctg)

```

(2)hdlc

```{r}
frmgham%>% 
  select(randid,totchol_ctg, hdlc, cursmoke) %>% 
  mutate(hdlc_ctg = cut(hdlc,breaks=c(-Inf, 40, 60, Inf), labels = c("Low", "Normal","Good"))) %>% 
  na.omit() %>% 
  group_by(totchol_ctg,hdlc_ctg) %>% 
  dplyr::summarise(percent.currentsmoke = sum(cursmoke)/n(),
                   n = n())

frmgham %>%
  select(randid,totchol_ctg, hdlc, cursmoke) %>% 
  mutate(hdlc_ctg = cut(hdlc,breaks=c(-Inf, 40, 60, Inf), labels = c("Low", "Normal","Good"))) %>% 
  na.omit() %>%  
  group_by(totchol_ctg,hdlc_ctg) %>%
  summarise(num.currentsmoke = sum(cursmoke)) %>% 
  ggplot(., aes(as.factor(totchol_ctg),num.currentsmoke)) +
    geom_bar(stat = "identity")+
  facet_wrap(~hdlc_ctg)
```


(3)ldlc

```{r}
frmgham%>% 
  select(randid,totchol_ctg, ldlc, cursmoke) %>% 
  mutate(ldlc_ctg = ifelse(ldlc>=130, "High","Ideal")) %>% 
  na.omit() %>% 
  group_by(totchol_ctg,ldlc_ctg) %>% 
  dplyr::summarise(percent.currentsmoke = sum(cursmoke)/n(),
                   n = n())


frmgham %>%
  select(randid,totchol_ctg, ldlc, cursmoke) %>% 
  mutate(ldlc_ctg = ifelse(ldlc>=130, "High","Ideal")) %>% 
  na.omit()  %>%  
  group_by(totchol_ctg,ldlc_ctg) %>%
  summarise(num.currentsmoke = sum(cursmoke)) %>% 
  ggplot(., aes(as.factor(totchol_ctg),num.currentsmoke)) +
    geom_bar(stat = "identity")+
  facet_wrap(~ldlc_ctg)
```

From the table above, we found age, ldlc and hdlc are confounders. 

## model fitting


```{r}
library(lme4)
cor(frmgham[,c("totchol", "age", "hdlc", "ldlc")],use = "pairwise.complete.obs")
m1 = glmer(cursmoke ~ totchol+ age+ hdlc+ldlc+(1|randid),
      data = frmgham, nAGQ=0,
      family = binomial(link=logit))
summary(m1)
se <- sqrt(diag(vcov(m1)))
# table of estimates with 95% CI
(tab <- cbind(Est = fixef(m1), LL = fixef(m1) - 1.96 * se, UL = fixef(m1) + 1.96 *
    se))
exp(tab)

```



