---
title: "logitudinal_final_proj"
author: "Lynette Pan"
date: "12/3/2018"
output: pdf_document
---

#Final Project: Smoking in the Framingham Heart Study


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(janitor)
library(mice)
library(gee)
library(lme4)

frmgham = 
  read.csv2("frmgham2.csv", sep = ",", na.strings=c("","NA")) %>% 
  clean_names() %>% select(cursmoke, everything()) %>% 
  mutate(sex = as.factor(sex),
         bpmeds = as.factor(bpmeds),
         educ = as.factor(educ),
         diabetes = as.factor(diabetes),
         prevap = as.factor(prevap),
         prevchd = as.factor(prevchd),
         prevmi = as.factor(prevmi),
         prevstrk = as.factor(prevstrk),
         prevhyp = as.factor(prevhyp),
         sysbp = as.numeric(levels(sysbp))[sysbp],
         diabp = as.numeric(as.character(diabp)),
         bmi = as.numeric(as.character(bmi))) %>% 
  as.tibble()

## created age category variable (age_ctg)
frmgham =
  frmgham %>% 
  mutate(
      age_ctg = case_when(
      age < 35 ~ "35-",  
      age %in% 35:44 ~ "35-44",
      age %in% 45:54 ~ "45-54",
      age %in% 55:64 ~ "55-64",
      age %in% 65:74 ~ "65-74",
      age >= 75 ~ "75+"
    ),
    age_ctg = fct_relevel(age_ctg, "35-")) 

## created sysbp category variable (sysbp_ctg)
### According to the new guideline from american heart association
### https://www.ahajournals.org/doi/10.1161/HYP.0000000000000065
frmgham =
  frmgham %>% 
  mutate(
      sysbp_ctg = case_when(
      sysbp < 120 ~ "normal",  
      (sysbp >= 120 & sysbp < 130) ~ "elevated",
      (sysbp >= 130 & sysbp < 140) ~ "stage1",
      (sysbp >= 140 & sysbp <= 180) ~ "stage2",
      sysbp > 180 ~ "crisis"
    ),
    sysbp_ctg = fct_relevel(sysbp_ctg, "normal")) 

## created sysbp category variable (diabp_ctg)
### According to the new guideline from american heart association
### https://www.ahajournals.org/doi/10.1161/HYP.0000000000000065
frmgham =
  frmgham %>% 
  mutate(
      diabp_ctg = case_when(
      diabp < 80 ~ "normal and elevated",  
      (diabp >= 80 & diabp < 90) ~ "stage1",
      (diabp >= 90 & diabp <= 120 ) ~ "stage2",
      diabp > 120 ~ "crisis"
      ),
    diabp_ctg = fct_relevel(diabp_ctg, "normal and elevated")) 

# select(frmgham, -randid) %>% 
#   lapply(., as.numeric)%>% 
#   cor(.)
```

# Descriptive statistics
```{r}
#descriptive statistics for the data

summary(frmgham)
```



#missing values
```{r}
# frmgham %>% 
#   group_by(period) %>% 
#   summarize(sum_na_period = sum(is.na(period)))
# 
# d1 = frmgham %>% 
#   group_by(randid) %>% 
#   summarize(n = n())
# table(d1$n)
# # 447 patients have only one obs
# # 781 patients have only two obs
# # 3206 patients have all three obs
# 
# number.subjects <- 
#   frmgham  %>% 
#   select(randid) %>% 
#   distinct() %>% 
#   unlist() %>%
#   length()
# 
# number.subjects
# 
# # missing data pattern by variables
# 
# sort(sapply(frmgham, function(x) { sum(is.na(x)) }), decreasing=TRUE)

```

# Impute missing values using MICE package
```{r}
# load("frmgham_complete.RData")
# # frmgham_temp <- mice(frmgham,m=1, method='cart', printFlag=FALSE,seed = 1)
# # frmgham
# # summary(frmgham_temp)
# # frmgham_complete <- complete(frmgham_temp)
# #save(frmgham_complete, file = "frmgham_complete.RData")
# frmgham_complete =
#   frmgham_complete %>% 
#   mutate(
#       age_ctg = case_when(
#       age < 35 ~ "35-",  
#       age %in% 35:44 ~ "35-44",
#       age %in% 45:54 ~ "45-54",
#       age %in% 55:64 ~ "55-64",
#       age %in% 65:74 ~ "65-74",
#       age >= 75 ~ "75+"
#     )) 
# 
# sort(sapply(frmgham_complete, function(x) { sum(is.na(x)) }), decreasing=TRUE)

```


* The output tells us that 2243 samples are complete, 7077 samples miss both hdlc and ldlc, 4 samples miss only the glucose value and so on.


* there are 447 patients who have only one obs; there are 781 patients who have only two obs and 3206 patients have all three obs. In total, there are 4434 patients in the study.

##1. Target question:
1) Is there a relationship between age and smoking status?  Does this relationship differ by sex? 

### cursmoke vs age_ctg 
```{r}
frmgham %>% 
  select(randid, cursmoke, age_ctg) %>% 
  na.omit() %>% 
  group_by(age_ctg) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n()) %>% 
  ggplot(aes(x = age_ctg, y = p_cursmoke)) + 
  geom_bar(stat = "identity") +
  labs(title = "Relationship between categorized age and smoking status",
       x = "Age",
       y = "Probability of current smoking")
```

### cursmoke vs age_ctg by sex 
```{r}
frmgham %>% 
  mutate(sex = ifelse(sex == 1, "male", "women")) %>%
  select(randid, cursmoke, age_ctg, sex) %>% 
  na.omit() %>% 
  group_by(age_ctg, sex) %>% 
  summarise(p_cursmoke = sum(cursmoke)/n()) %>% 
  ggplot(aes(x = age_ctg, y = p_cursmoke, color = sex, group = sex)) + 
  geom_line() +
  geom_point() +
  labs(title = "Relationship between categorized age and smoking status by sex",
       x = "Age",
       y = "Probability of current smoking")
```


### See random effects 


## Finding confounders
### age and cursmoke -- confounder: hdlc, ldlc
```{r}
age <- 
  gee(cursmoke ~ age, 
      data = frmgham, 
      id = randid, 
      family = binomial,
      corstr = ("unstructured"))

pot_conf <- names(frmgham)[-c(1:3, 5)]
pot_conf <- 
  pot_conf[-c(grep("time", pot_conf), frmgham$period, frmgham$death)]

conf_det <- vector("list", length = length(pot_conf))
names(conf_det) <- pot_conf



for (conf in pot_conf) {
  
  data <- cbind(frmgham[,c("cursmoke", "age")], frmgham[,conf])
  try(fit_gee <- gee(cursmoke ~ ., 
      data = data, 
      id = frmgham$randid, 
      family = binomial,
      corstr = ("unstructured")))
 
  result <- tibble(
    variable = c("age"),
    OR =
      exp(coef(summary(age))[2,1]),
    lower_CI  =
      exp(coef(summary(age))[2,1] - 1.96*(coef(summary(age))[2,4])),
    upper_CI =
      exp(coef(summary(age))[2,1] + 1.96*(coef(summary(age))[2,4])),
    potconf_OR =
      exp(coef(summary(fit_gee))[2,1]),
    confounder =
      if_else((potconf_OR >= lower_CI & potconf_OR <= upper_CI), FALSE, TRUE)
  )

  conf_det[[conf]] <- result
}

map(conf_det, knitr::kable, digits = 3)
```





(2) Is there a relationship between the number of cigarettes smoked per day and age?  Does this relationship differ by sex? 

While answering these questions, please account for any confounders that you have evidence may impact the relationship between age and sex with smoking.

```{r}
frmgham %>% 
  ggplot(., aes(x=age, y=cigpday)) +
    geom_jitter() + 
    geom_smooth(method = "loess") +
  ggtitle("Unadjusted relationship between the number of cigarettes smoked per day and age")
```

```{r finding counfounders for m2}
# I will look into SYSBP DIABP TOTCHOL BMI DIABETES 
# based on https://tobaccocontrol.bmj.com/content/14/5/315

cigpday.age <- 
  gee(cigpday ~ age, 
      data = frmgham, 
      id = randid, 
      family = poisson,
      corstr = ("unstructured"))

pot_conf <- names(frmgham)[-c(2, 3, 5, 8)]
pot_conf <- 
  pot_conf[-c(grep("time", pot_conf))]

conf_det <- vector("list", length = length(pot_conf))
names(conf_det) <- pot_conf


for (conf in pot_conf) {
  
  data <- cbind(frmgham[,c("cigpday", "age")], frmgham[,conf])
  try(fit_gee <- gee(cigpday ~ ., 
      data = data, 
      id = frmgham$randid, 
      family = poisson,
      corstr = ("unstructured")))
 
  result <- tibble(
    variable = c("age"),
    OR =
      exp(coef(summary(cigpday.age))[2,1]),
    lower_CI  =
      exp(coef(summary(cigpday.age))[2,1] - 1.96*(coef(summary(cigpday.age))[2,4])),
    upper_CI =
      exp(coef(summary(cigpday.age))[2,1] + 1.96*(coef(summary(cigpday.age))[2,4])),
    potconf_OR =
      exp(coef(summary(fit_gee))[2,1]),
    confounder =
      if_else((potconf_OR >= lower_CI & potconf_OR <= upper_CI), FALSE, TRUE)
  )

  conf_det[[conf]] <- result
}

map(conf_det, knitr::kable, digits = 3)
```



```{r}
frmgham %>% 
  mutate(sex=ifelse(sex==1, "male","women")) %>% 
  ggplot(., aes(x=age, y=cigpday)) +
  geom_jitter() + 
  geom_smooth(method = "loess") +
  facet_grid(~sex)+
  ggtitle("Relationship between the number of cigarettes smoked per day and age by gender")
```




#2.Target question:

Next you are interested in the relationship between certain health outcomes and smoking status. In particular you are interested in :


(1)The relationship between current smoking status and systolic blood pressure.
```{r}
frmgham %>% 
  ggplot(., aes(x=sysbp, y=cursmoke)) +
    geom_jitter() + 
    geom_smooth(method = "loess") + 
  ggtitle("Unadjusted relationship between current smoking status and diastolic blood pressure")

#https://www.ncbi.nlm.nih.gov/pubmed/7790985
#From the research paper above, diastolic blood pressure and hypertension are potential confounders.

#center the variables and find the confounder
frmgham$diabp_c = frmgham$diabp-mean(frmgham$diabp)
frmgham$sysbp_c = frmgham$sysbp-mean(frmgham$sysbp)

sysbp <- 
  glmer(cursmoke ~ sysbp_c + (1|randid), 
      data = frmgham, 
      family = binomial)

sysbp_diabp <- 
  glmer(cursmoke ~ sysbp_c + diabp_c + (1|randid), 
      data = frmgham, 
      family = binomial)

sysbp_prevhyp <- 
  glmer(cursmoke ~ sysbp_c + prevhyp + (1|randid), 
      data = frmgham, 
      family = binomial)

sysbp_hyperten <- 
  glmer(cursmoke ~ sysbp_c + hyperten+ (1|randid), 
      data = frmgham, 
      family = binomial)

tibble(
    variable = c("diabp", "prevhyp", "hyperten"),
    OR = 
      exp(coef(summary(sysbp))[2,1]),
    lower_CI  = 
      exp(coef(summary(sysbp))[2,1] - 1.96*(coef(summary(sysbp))[2,4])),
    upper_CI = 
      exp(coef(summary(sysbp))[2,1] + 1.96*(coef(summary(sysbp))[2,4])),
    conf_OR =
      c(exp(coef(summary(sysbp_diabp))[2,1]),
        exp(coef(summary(sysbp_prevhyp))[2,1]),
        exp(coef(summary(sysbp_hyperten))[2,1])),
    confounder =
      if_else((conf_OR >= lower_CI & conf_OR <= upper_CI), FALSE, TRUE)
  )

smoke_diabp = glmer(cursmoke ~ diabp_c + sysbp_c + prevhyp + hyperten+ (1|randid), 
      data = frmgham, 
      family = binomial)
summary(smoke_diabp)
```

(2)The relationship between current smoking status and diastolic blood pressure.
```{r}
frmgham %>% 
  ggplot(., aes(x=diabp, y=cursmoke)) +
    geom_jitter() + 
    geom_smooth(method = "loess") + 
  ggtitle("Unadjusted relationship between current smoking status and diastolic blood pressure")

#https://www.ncbi.nlm.nih.gov/pubmed/7790985
#From the research paper above, systolic blood pressure and hypertension are potential confounders.

#center the variables and find the confounder
frmgham$diabp_c = frmgham$diabp-mean(frmgham$diabp)
frmgham$sysbp_c = frmgham$sysbp-mean(frmgham$sysbp)

diabp <- 
  glmer(cursmoke ~ diabp_c + (1|randid), 
      data = frmgham, 
      family = binomial)

diabp_sysbp <- 
  glmer(cursmoke ~ diabp_c + sysbp_c + (1|randid), 
      data = frmgham, 
      family = binomial)

diabp_prevhyp <- 
  glmer(cursmoke ~ diabp_c + prevhyp + (1|randid), 
      data = frmgham, 
      family = binomial)

diabp_hyperten <- 
  glmer(cursmoke ~ diabp_c + hyperten+ (1|randid), 
      data = frmgham, 
      family = binomial)

tibble(
    variable = c("sysbp", "prevhyp", "hyperten"),
    OR = 
      exp(coef(summary(diabp))[2,1]),
    lower_CI  = 
      exp(coef(summary(diabp))[2,1] - 1.96*(coef(summary(diabp))[2,4])),
    upper_CI = 
      exp(coef(summary(diabp))[2,1] + 1.96*(coef(summary(diabp))[2,4])),
    conf_OR =
      c(exp(coef(summary(diabp_sysbp))[2,1]),
        exp(coef(summary(diabp_prevhyp))[2,1]),
        exp(coef(summary(diabp_hyperten))[2,1])),
    confounder =
      if_else((conf_OR >= lower_CI & conf_OR <= upper_CI), FALSE, TRUE)
  )

smoke_diabp = glmer(cursmoke ~ diabp_c + sysbp_c + prevhyp + hyperten+ (1|randid), 
      data = frmgham, 
      family = binomial)
summary(smoke_diabp)

#confounders: sysbp, prevhyp, hyperten
```

From the table above, we found "sysbp", "prevhyp" and "hyperten" are confounders. After adding the confounders, the final model is shown above. For a one-unit increase in the diastolic blood pressure, the expected change in log odds is 0.009, adjusting for systolic blood pressure, prevalent hypertensive and hypertensive.

(3)The relationship between current smoking status and serum total cholesterol. Again, while answering these questions, please account for any confounders that you have evidence may impact these relationships. 


```{r}
frmgham %>% 
  ggplot(., aes(x=totchol, y=cursmoke)) +
    geom_jitter() + 
    geom_smooth(method = "loess") +ggtitle("Unadjusted relationship between current smoking status and serum total cholesterol")


#https://medlineplus.gov/cholesterollevelswhatyouneedtoknow.html
#categorized totchol based on link above

frmgham=frmgham %>% 
  mutate(totchol_ctg= ifelse(age<=19 & totchol<170,0,
                             ifelse(age>19&totchol %in% c(125, 200), 0,1)))
table(frmgham$totchol_ctg)
# 124 normal total chol 
# 11503 abnormal total chol
```

## Finding confounders

(a) Sex -should be adjusted
```{r}
frmgham %>% 
  ggplot(., aes(x=totchol, y=cursmoke)) +
    geom_jitter() + 
    geom_smooth(method = "loess") +ggtitle("Unadjusted relationship between current smoking status and serum total cholesterol")+
  facet_grid(~sex)

null_model=gee(cursmoke ~ totchol, data = frmgham,id = randid,family = binomial,corstr = ("unstructured"))
summary(null_model)
#without adjusting for sex
null_ci=c(exp(coef(summary(null_model))[2, 1]+ 1.96*coef(summary(null_model))[2, 4]),exp(coef(summary(null_model))[2, 1]- 1.96*coef(summary(null_model))[2, 4]))
#[1] 0.9998513 0.9981607
#adjusting for sex
sex_totchol=gee(cursmoke ~ totchol+sex, data = frmgham,id = randid,family = binomial,corstr = ("unstructured"))
summary(sex_totchol)
exp(coef(summary(sex_totchol))[2, 1]) %in% null_ci

frmgham%>% 
  select(randid,totchol_ctg, sex, cursmoke) %>% 
  na.omit() %>% 
  group_by(sex,totchol_ctg) %>% 
  dplyr::summarise(percent.currentsmoke = sum(cursmoke)/n(),
                   n = n())

```

* Sex does confound the relationship between between current smoking status and serum total cholesterol since 0.9994668 is in the 95% CI from the model without sex adjusted.


(b) PREVCHD 

```{r}
frmgham %>% 
  ggplot(., aes(x=totchol, y=cursmoke)) +
    geom_jitter() + 
    geom_smooth(method = "loess") +ggtitle("Unadjusted relationship between current smoking status and serum total cholesterol")+
  facet_grid(~prevchd)

#adjusting for prevchd
prevchd_totchol=gee(cursmoke ~ totchol+prevchd, data = frmgham,id = randid,family = binomial,corstr = ("unstructured"))
summary(prevchd_totchol)
exp(coef(summary(prevchd_totchol))[2, 1]) %in% null_ci

frmgham%>% 
  select(randid,totchol_ctg, prevchd, cursmoke) %>% 
  na.omit() %>% 
  group_by(prevchd,totchol_ctg) %>% 
  dplyr::summarise(percent.currentsmoke = sum(cursmoke)/n(),
                   n = n())

```


* prevchd_totchol does confound the relationship between between current smoking status and serum total cholesterol.

(3) BMI
```{r}
frmgham= frmgham %>% 
  mutate(bmi=as.numeric(as.character(bmi))) %>% 
  mutate(bmi_cat = ifelse(bmi<18.5,0,ifelse(bmi %in% c(18.5,24.9), 1, ifelse(bmi %in% c(25, 29.9),2, 3))))
#adjusting for prevchd
bmi_totchol=gee(cursmoke ~ totchol+bmi, data = frmgham,id = randid,family = binomial,corstr = ("unstructured"))
summary(bmi_totchol)
exp(coef(summary(bmi_totchol))[2, 1]) %in% null_ci #bmi as numeric not confounder

bmi_cat_totchol=gee(cursmoke ~ totchol+bmi_cat, data = frmgham,id = randid,family = binomial,corstr = ("unstructured"))
exp(coef(summary(bmi_cat_totchol))[2, 1]) %in% null_ci


frmgham%>% 
  select(randid,totchol_ctg, bmi_cat, cursmoke) %>% 
  na.omit() %>% 
  group_by(bmi_cat,totchol_ctg) %>% 
  dplyr::summarise(percent.currentsmoke = sum(cursmoke)/n(),
                   n = n())
```

* bmi_cat and bmi do confound the relationship between between current smoking status and serum total cholesterol.


(3) age


```{r}
#adjusting for age
age_totchol=gee(cursmoke ~ totchol+age, data = frmgham,id = randid,family = binomial,corstr = ("unstructured"))
summary(age_totchol)
exp(coef(summary(age_totchol))[2, 1]) %in% null_ci #age as numeric not confounder


frmgham%>% 
  select(randid,totchol_ctg, age_ctg, cursmoke) %>% 
  na.omit() %>% 
  group_by(age_ctg,totchol_ctg) %>% 
  dplyr::summarise(percent.currentsmoke = sum(cursmoke)/n(),
                   n = n())
```

(5) prevap


```{r}
prevap_totchol=gee(cursmoke ~ totchol+prevap, data = frmgham,id = randid,family = binomial,corstr = ("unstructured"))
summary(prevap_totchol)
exp(coef(summary(prevap_totchol))[2, 1]) %in% null_ci #prevap is a confounder


frmgham%>% 
  select(randid,totchol_ctg, prevap, cursmoke) %>% 
  na.omit() %>% 
  group_by(prevap,totchol_ctg) %>% 
  dplyr::summarise(percent.currentsmoke = sum(cursmoke)/n(),
                   n = n())
```

(6) prevmi
```{r}

prevmi_totchol=gee(cursmoke ~ totchol+prevmi, data = frmgham,id = randid,family = binomial,corstr = ("unstructured"))
summary(prevmi_totchol)
exp(coef(summary(prevmi_totchol))[2, 1]) %in% null_ci #prevmi is a confounder


frmgham%>% 
  select(randid,totchol_ctg, prevmi, cursmoke) %>% 
  na.omit() %>% 
  group_by(prevmi,totchol_ctg) %>% 
  dplyr::summarise(percent.currentsmoke = sum(cursmoke)/n(),
                   n = n())
```

(7) storke
```{r}

prevstrk_totchol=gee(cursmoke ~ totchol+prevstrk, data = frmgham,id = randid,family = binomial,corstr = ("unstructured"))
summary(prevstrk_totchol)
exp(coef(summary(prevstrk_totchol))[2, 1]) %in% null_ci #prevmi is a confounder


frmgham%>% 
  select(randid,totchol_ctg, prevstrk, cursmoke) %>% 
  na.omit() %>% 
  group_by(prevstrk,totchol_ctg) %>% 
  dplyr::summarise(percent.currentsmoke = sum(cursmoke)/n(),
                   n = n())
```
>>>>>>> b25b10657ce697ce984e8dedae8d96ec307a2c04
